---
title: "Deserts x Race/Ethnicity: Asian American/Pacific Islander and American Indian/Alaska Native Communities"
author: "Hailey Gibbs"
date: "June 16, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

This file runs two **independent, parallel** classification systems against
the same childcare desert data:

1. **AAPI** = Asian alone (P8_006N) + Native Hawaiian and Other Pacific
   Islander alone (P8_008N), summed as one combined share, with separate
   breakdowns also shown for Asian alone and NHPI alone individually.
2. **AIAN** = American Indian and Alaska Native alone (P8_005N), tracked on
   its own — not combined with AAPI in any shared sum.

Both 50% ("majority") and 25% ("predominantly") thresholds are calculated for
each system. Aggregation is at the county level, and rurality (from
`joined_rural`) is incorporated according to previous analyses on
race/ethnicity: a rural flag joined on longitude/latitude, with desert rates
compared by composition × rurality, and state-level driver tables limited to
rural areas.

```{r}
library(tidyverse)
desert_threshold <- 0.33

# Loading csv files
aapi <- read_csv("DECENNIALDHC2020.P8-Data.csv", col_names = TRUE, show_col_types = FALSE)

# Sanity check: confirm the file loaded as expected before proceeding.
# The P8 table should have ~73-74 columns (GEO_ID, NAME, P8_001N...P8_071N)
# and GEO_ID should contain "1400000US" tract identifiers, not NAs or an
# unexpected number of columns. If this check fails, STOP — do not proceed
# with downstream analysis until the correct file is confirmed loaded.
cat("Columns loaded:", ncol(aapi), "\n")
cat("Column names (first 10):", paste(names(aapi)[1:10], collapse = ", "), "\n")
cat("First GEO_ID value:", aapi$GEO_ID[1], "\n")
cat("Second GEO_ID value (first data row):", aapi$GEO_ID[2], "\n")

if (!"GEO_ID" %in% names(aapi)) {
  stop("GEO_ID column not found — wrong file may have loaded. Check filename and re-run.")
}
if (!str_detect(aapi$GEO_ID[2], "^1400000US")) {
  stop("Second row of GEO_ID does not match expected tract format '1400000US...' — wrong file may have loaded.")
}

aapi <- aapi[-1, ]   # row 1 is the variable-label row, not data

childcare_data <- read_csv("cap_ccaccess_2025_revised_02282026.csv")
joined_rural   <- read_csv("joined_rural.csv")
```

## Data cleaning

```{r}
# Cleaning the race data
# P8_001N = Total population
# P8_005N = American Indian and Alaska Native alone
# P8_006N = Asian alone
# P8_008N = Native Hawaiian and Other Pacific Islander alone
race_data <- aapi %>%
  mutate(
    total_pop   = as.numeric(P8_001N),
    aian_alone  = as.numeric(P8_005N),
    asian_alone = as.numeric(P8_006N),
    nhpi_alone  = as.numeric(P8_008N),

    pct_aian  = if_else(total_pop > 0, (aian_alone  / total_pop) * 100, NA_real_),
    pct_asian = if_else(total_pop > 0, (asian_alone / total_pop) * 100, NA_real_),
    pct_nhpi  = if_else(total_pop > 0, (nhpi_alone  / total_pop) * 100, NA_real_),
    pct_aapi  = if_else(total_pop > 0, ((asian_alone + nhpi_alone) / total_pop) * 100, NA_real_),

    # Strip the "1400000US" prefix to get the 11-digit census tract FIPS
    county_fips = str_sub(str_remove(GEO_ID, "1400000US"), 1, 5)
  ) %>%
  filter(!is.na(pct_aian), !is.na(pct_asian), !is.na(pct_nhpi), !is.na(pct_aapi)) %>%
  group_by(county_fips) %>%
  summarise(
    avg_pct_aian  = weighted.mean(pct_aian,  w = total_pop, na.rm = TRUE),
    avg_pct_asian = weighted.mean(pct_asian, w = total_pop, na.rm = TRUE),
    avg_pct_nhpi  = weighted.mean(pct_nhpi,  w = total_pop, na.rm = TRUE),
    avg_pct_aapi  = weighted.mean(pct_aapi,  w = total_pop, na.rm = TRUE),
    .groups = "drop"
  )

# Sanity check on race_data before the join: percentages should be bounded
# 0-100 and we should have one row per county (not per tract)
cat("Number of counties in race_data:", nrow(race_data), "\n")
cat("Range of avg_pct_aapi:", round(range(race_data$avg_pct_aapi), 2), "\n")
cat("Range of avg_pct_aian:", round(range(race_data$avg_pct_aian), 2), "\n")
if (any(race_data$avg_pct_aapi > 100, na.rm = TRUE) || any(race_data$avg_pct_aian > 100, na.rm = TRUE)) {
  stop("A county-level percentage exceeds 100% — check upstream arithmetic before proceeding.")
}

# Joining with childcare data
childcare_with_race <- childcare_data %>%
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  left_join(race_data, by = "county_fips") %>%
  mutate(desert = adj_supply <= desert_threshold)

# Check join quality
cat("Rows with AAPI data:", sum(!is.na(childcare_with_race$avg_pct_aapi)), "\n")
cat("Rows missing AAPI data:", sum(is.na(childcare_with_race$avg_pct_aapi)), "\n")
cat("Rows with AIAN data:", sum(!is.na(childcare_with_race$avg_pct_aian)), "\n")
cat("Rows missing AIAN data:", sum(is.na(childcare_with_race$avg_pct_aian)), "\n")

# Create rural flag
rural_lookup <- joined_rural %>%
  select(longitude, latitude, rural) %>%
  distinct()

# Joining with rural flag
childcare_with_race_rural <- childcare_with_race %>%
  left_join(rural_lookup, by = c("longitude", "latitude"))

# Check join quality
cat("Rows with rural status:", sum(!is.na(childcare_with_race_rural$rural)), "\n")
cat("Rows missing rural status:", sum(is.na(childcare_with_race_rural$rural)), "\n")

# Remove missing values and code new variables.
# Two independent classification systems (AAPI, AIAN), each at two
# thresholds (50%, 25%) — four county_type columns total.
childcare_with_race_rural <- childcare_with_race_rural %>%
  filter(
    !is.na(adj_supply),
    !is.na(avg_pct_asian), !is.na(avg_pct_nhpi), !is.na(avg_pct_aapi),
    !is.na(avg_pct_aian),
    !is.na(rural)
  ) %>%
  mutate(
    desert = adj_supply <= desert_threshold,

    # ---- AAPI: 50% ("majority") threshold ----
    county_type_aapi_50 = case_when(
      avg_pct_asian >= 50 ~ "Majority Asian",
      avg_pct_nhpi  >= 50 ~ "Majority NHPI",
      avg_pct_aapi  >= 50 ~ "Majority AAPI (combined)",
      TRUE                 ~ "Not majority AAPI"
    ),

    # ---- AAPI: 25% ("predominantly") threshold ----
    county_type_aapi_25 = case_when(
      avg_pct_asian >= 25 ~ "Predominantly Asian",
      avg_pct_nhpi  >= 25 ~ "Predominantly NHPI",
      avg_pct_aapi  >= 25 ~ "Predominantly AAPI (combined)",
      TRUE                 ~ "Not predominantly AAPI"
    ),

    # ---- AIAN: 50% ("majority") threshold — independent of AAPI ----
    county_type_aian_50 = case_when(
      avg_pct_aian >= 50 ~ "Majority AIAN",
      TRUE                ~ "Not majority AIAN"
    ),

    # ---- AIAN: 25% ("predominantly") threshold — independent of AAPI ----
    county_type_aian_25 = case_when(
      avg_pct_aian >= 25 ~ "Predominantly AIAN",
      TRUE                ~ "Not predominantly AIAN"
    )
  )

# Retain the non-rural-filtered version for analyses that don't need rural
# status (kept under the original name for backward compatibility)
childcare_with_race <- childcare_with_race %>%
  filter(
    !is.na(adj_supply),
    !is.na(avg_pct_asian), !is.na(avg_pct_nhpi), !is.na(avg_pct_aapi),
    !is.na(avg_pct_aian)
  ) %>%
  mutate(
    desert = adj_supply <= desert_threshold,
    county_type_aapi_50 = case_when(
      avg_pct_asian >= 50 ~ "Majority Asian",
      avg_pct_nhpi  >= 50 ~ "Majority NHPI",
      avg_pct_aapi  >= 50 ~ "Majority AAPI (combined)",
      TRUE                 ~ "Not majority AAPI"
    ),
    county_type_aapi_25 = case_when(
      avg_pct_asian >= 25 ~ "Predominantly Asian",
      avg_pct_nhpi  >= 25 ~ "Predominantly NHPI",
      avg_pct_aapi  >= 25 ~ "Predominantly AAPI (combined)",
      TRUE                 ~ "Not predominantly AAPI"
    ),
    county_type_aian_50 = case_when(
      avg_pct_aian >= 50 ~ "Majority AIAN",
      TRUE                ~ "Not majority AIAN"
    ),
    county_type_aian_25 = case_when(
      avg_pct_aian >= 25 ~ "Predominantly AIAN",
      TRUE                ~ "Not predominantly AIAN"
    )
  )
```

## State aggregated data

```{r}
# Creating state-level dataset (AAPI and AIAN shares shown side by side
# for reference, but classification systems below remain independent)
state_race_desert <- childcare_with_race %>%
  filter(!is.na(avg_pct_aapi), !is.na(avg_pct_aian), !is.na(adj_supply)) %>%
  group_by(state_name) %>%
  summarise(
    pct_in_desert  = round(mean(desert) * 100, 1),
    avg_pct_asian  = round(mean(avg_pct_asian), 1),
    avg_pct_nhpi   = round(mean(avg_pct_nhpi), 1),
    avg_pct_aapi   = round(mean(avg_pct_aapi), 1),
    avg_pct_aian   = round(mean(avg_pct_aian), 1),
    .groups = "drop"
  )

print(state_race_desert)
```

## AAPI: Comparing desert rates by county composition — 50% threshold

```{r}
childcare_with_race %>%
  group_by(county_type_aapi_50) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))
```

## AAPI: Comparing desert rates by county composition — 25% threshold

```{r}
childcare_with_race %>%
  group_by(county_type_aapi_25) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))
```

## AIAN: Comparing desert rates by county composition — 50% threshold

```{r}
childcare_with_race %>%
  group_by(county_type_aian_50) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))
```

## AIAN: Comparing desert rates by county composition — 25% threshold

```{r}
childcare_with_race %>%
  group_by(county_type_aian_25) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))
```

## AAPI: Comparing desert rates by county composition and urbanicity

```{r}
# 50% threshold x rurality
childcare_with_race_rural %>%
  group_by(county_type_aapi_50, rural) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(county_type_aapi_50, rural)

# 25% threshold x rurality
childcare_with_race_rural %>%
  group_by(county_type_aapi_25, rural) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(county_type_aapi_25, rural)
```

## AIAN: Comparing desert rates by county composition and urbanicity

```{r}
# 50% threshold x rurality
childcare_with_race_rural %>%
  group_by(county_type_aian_50, rural) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(county_type_aian_50, rural)

# 25% threshold x rurality
childcare_with_race_rural %>%
  group_by(county_type_aian_25, rural) %>%
  summarise(
    total_areas   = n(),
    pct_in_desert = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(county_type_aian_25, rural)
```

## AAPI: States driving majority/predominantly AAPI desert rates

```{r}
# 50% threshold, by state — includes distinct county count for context
state_aapi_50 <- childcare_with_race %>%
  filter(county_type_aapi_50 != "Not majority AAPI") %>%
  group_by(state_name, county_type_aapi_50) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aapi_50)

# 25% threshold, by state
state_aapi_25 <- childcare_with_race %>%
  filter(county_type_aapi_25 != "Not predominantly AAPI") %>%
  group_by(state_name, county_type_aapi_25) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aapi_25)
```

## AIAN: States driving majority/predominantly AIAN desert rates

```{r}
# 50% threshold, by state
state_aian_50 <- childcare_with_race %>%
  filter(county_type_aian_50 == "Majority AIAN") %>%
  group_by(state_name, county_type_aian_50) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aian_50)

# 25% threshold, by state
state_aian_25 <- childcare_with_race %>%
  filter(county_type_aian_25 == "Predominantly AIAN") %>%
  group_by(state_name, county_type_aian_25) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aian_25)
```

## States driving rural majority/predominantly AAPI and AIAN desert rates

```{r}
# Rural AAPI, 50% threshold, by state
state_aapi_rural_50 <- childcare_with_race_rural %>%
  filter(rural == 1, county_type_aapi_50 != "Not majority AAPI") %>%
  group_by(state_name, county_type_aapi_50) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aapi_rural_50)

# Rural AAPI, 25% threshold, by state
state_aapi_rural_25 <- childcare_with_race_rural %>%
  filter(rural == 1, county_type_aapi_25 != "Not predominantly AAPI") %>%
  group_by(state_name, county_type_aapi_25) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aapi_rural_25)

# Rural AIAN, 50% threshold, by state
state_aian_rural_50 <- childcare_with_race_rural %>%
  filter(rural == 1, county_type_aian_50 == "Majority AIAN") %>%
  group_by(state_name, county_type_aian_50) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aian_rural_50)

# Rural AIAN, 25% threshold, by state
state_aian_rural_25 <- childcare_with_race_rural %>%
  filter(rural == 1, county_type_aian_25 == "Predominantly AIAN") %>%
  group_by(state_name, county_type_aian_25) %>%
  summarise(
    total_areas         = n(),
    n_distinct_counties  = n_distinct(county_fips),
    pct_in_desert        = round(mean(desert) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

print(state_aian_rural_25)
```

## Diagnostic check: small-county artifacts

Following the same logic used to check the Kansas/Oklahoma/Puerto Rico NHPI
rows earlier, this chunk flags any state/county_type combination — across
**both** AAPI and AIAN systems — where the number of distinct counties is
small relative to total_areas, which can produce misleadingly extreme
pct_in_desert values driven by tiny denominators rather than real patterns.

```{r}
# Flag combinations backed by very few counties (adjust threshold as needed)
flag_small_n <- function(df, type_col) {
  df %>%
    filter(n_distinct_counties <= 3) %>%
    select(state_name, {{ type_col }}, total_areas, n_distinct_counties, pct_in_desert)
}

cat("AAPI 50% — small-county flags:\n")
print(flag_small_n(state_aapi_50, county_type_aapi_50))

cat("\nAAPI 25% — small-county flags:\n")
print(flag_small_n(state_aapi_25, county_type_aapi_25))

cat("\nAIAN 50% — small-county flags:\n")
print(flag_small_n(state_aian_50, county_type_aian_50))

cat("\nAIAN 25% — small-county flags:\n")
print(flag_small_n(state_aian_25, county_type_aian_25))

cat("\nAAPI rural 50% — small-county flags:\n")
print(flag_small_n(state_aapi_rural_50, county_type_aapi_50))

cat("\nAAPI rural 25% — small-county flags:\n")
print(flag_small_n(state_aapi_rural_25, county_type_aapi_25))

cat("\nAIAN rural 50% — small-county flags:\n")
print(flag_small_n(state_aian_rural_50, county_type_aian_50))

cat("\nAIAN rural 25% — small-county flags:\n")
print(flag_small_n(state_aian_rural_25, county_type_aian_25))
```

## Diagnostic check: Puerto Rico / Kansas / Oklahoma NHPI rows

These rows showed implausibly high NHPI shares in prior runs. Before citing
any NHPI finding for these locations, this chunk recomputes percentages
directly from the raw P8 data, bypassing the full pipeline, to confirm
whether the join and aggregation logic produced these numbers correctly or
whether they reflect a data or coding error.

```{r}
diagnostic_states <- c("Kansas", "Oklahoma", "Puerto Rico")

# Distinct counties driving each state's "Predominantly NHPI" classification
diagnostic_detail <- childcare_with_race %>%
  filter(state_name %in% diagnostic_states, county_type_aapi_25 == "Predominantly NHPI") %>%
  distinct(state_name, county_fips, avg_pct_nhpi, avg_pct_asian, avg_pct_aapi)

print(diagnostic_detail, n = 50)

# Recompute NHPI percentage directly from raw P8 data for just these counties
flagged_counties <- diagnostic_detail$county_fips

raw_tract_check <- aapi %>%
  mutate(
    total_pop   = as.numeric(P8_001N),
    nhpi_alone  = as.numeric(P8_008N),
    county_fips = str_sub(str_remove(GEO_ID, "1400000US"), 1, 5)
  ) %>%
  filter(county_fips %in% flagged_counties) %>%
  group_by(county_fips) %>%
  summarise(
    n_tracts            = n(),
    county_total_pop    = sum(total_pop, na.rm = TRUE),
    county_nhpi_alone    = sum(nhpi_alone, na.rm = TRUE),
    recomputed_pct_nhpi  = round(county_nhpi_alone / county_total_pop * 100, 2),
    .groups = "drop"
  ) %>%
  left_join(diagnostic_detail %>% distinct(county_fips, state_name), by = "county_fips") %>%
  arrange(state_name, desc(recomputed_pct_nhpi))

print(raw_tract_check, n = 50)

# How many childcare points sit in each flagged county?
points_per_county <- childcare_with_race %>%
  filter(county_fips %in% flagged_counties) %>%
  count(state_name, county_fips, name = "n_childcare_points") %>%
  arrange(state_name, desc(n_childcare_points))

print(points_per_county, n = 50)
```

## Save outputs

```{r}
write_csv(childcare_with_race,        "aapi_aian_childcare_with_race.csv")
write_csv(childcare_with_race_rural,  "aapi_aian_childcare_with_race_rural.csv")
write_csv(state_race_desert,          "aapi_aian_state_desert_rates.csv")
write_csv(state_aapi_50,              "aapi_state_majority_50.csv")
write_csv(state_aapi_25,              "aapi_state_majority_25.csv")
write_csv(state_aian_50,              "aian_state_majority_50.csv")
write_csv(state_aian_25,              "aian_state_majority_25.csv")
write_csv(state_aapi_rural_50,        "aapi_state_rural_majority_50.csv")
write_csv(state_aapi_rural_25,        "aapi_state_rural_majority_25.csv")
write_csv(state_aian_rural_50,        "aian_state_rural_majority_50.csv")
write_csv(state_aian_rural_25,        "aian_state_rural_majority_25.csv")
```
