---
title: "Deserts x Race/Ethnicity: Asian American/Pacific Islander and American Indian/Alaska Native Communities"
author: "Hailey Gibbs"
date: "June 16, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

This file estimates the share of young children living in a licensed childcare desert for four community types — Asian, NHPI, AAPI (combined), and AIAN — broken out by urban/rural status and nationwide estimates. Note: NHPI nationwide estimates are excluded from final output due to insufficient sample size (zero counties meet the 25% threshold at the national level); urban and rural NHPI estimates are retained.

Classification uses **non-exclusive 25% flags**: a county is included in a group if that racial/ethnic group makes up at least 25% of the county population. The same county can appear in more than one group, since they are listed non-exclusively.
This workaround is uses to account for overall low population counts among these racial groups, and the average percent of each racial group is based on a weighted mean within that county.

- **AAPI** = Asian alone (P8_006N) + Native Hawaiian and Other Pacific
  Islander alone (P8_008N); also evaluated separately for Asian alone and
  NHPI alone.
- **AIAN** = American Indian and Alaska Native alone (P8_005N); tracked
  independently.

The unit of analysis is **young children**: each lat/long point in
`childcare_data` represents 10 synthetic families with a child aged 5 and
under, derived from Census data.

Puerto Rico is excluded from all analysis tables due to unreliable join. 
Rural status is joined from `joined_rural` on longitude/latitude.

```{r load}
library(tidyverse)
desert_threshold <- 0.33

# P8 race/ethnicity data (2020 Decennial DHC)
p8_raw <- read_csv("DECENNIALDHC2020.P8-Data.csv", col_names = TRUE, show_col_types = FALSE)

# Sanity check: GEO_ID should be present and contain tract-level identifiers.
# Stop immediately if the wrong file loaded.
cat("Columns loaded:", ncol(p8_raw), "\n")
cat("First GEO_ID value:", p8_raw$GEO_ID[1], "\n")
cat("Second GEO_ID value (first data row):", p8_raw$GEO_ID[2], "\n")

if (!"GEO_ID" %in% names(p8_raw)) {
  stop("GEO_ID column not found — wrong file may have loaded.")
}
if (!str_detect(p8_raw$GEO_ID[2], "^1400000US")) {
  stop("GEO_ID does not match expected tract format '1400000US...' — wrong file may have loaded.")
}

p8_raw        <- p8_raw[-1, ]   # row 1 is the variable-label row, not data
childcare_raw <- read_csv("cap_ccaccess_2025_revised_02282026.csv")
joined_rural  <- read_csv("joined_rural.csv")
```

## Data cleaning

```{r clean}
# --- Step 1: Aggregate P8 race shares from tract to county ---
# P8_001N = Total population
# P8_005N = American Indian and Alaska Native alone
# P8_006N = Asian alone
# P8_007N = Native Hawaiian and Other Pacific Islander alone

race_county <- p8_raw %>%
  mutate(
    total_pop   = as.numeric(P8_001N),
    aian_alone  = as.numeric(P8_005N),
    asian_alone = as.numeric(P8_006N),
    nhpi_alone  = as.numeric(P8_007N),
    pct_aian    = if_else(total_pop > 0, aian_alone  / total_pop * 100, NA_real_),
    pct_asian   = if_else(total_pop > 0, asian_alone / total_pop * 100, NA_real_),
    pct_nhpi    = if_else(total_pop > 0, nhpi_alone  / total_pop * 100, NA_real_),
    pct_aapi    = if_else(total_pop > 0, (asian_alone + nhpi_alone) / total_pop * 100, NA_real_),
    county_fips = str_sub(str_remove(GEO_ID, "1400000US"), 1, 5)
  ) %>%
  filter(!is.na(pct_aian), !is.na(pct_asian), !is.na(pct_nhpi), !is.na(pct_aapi)) %>%
  group_by(county_fips) %>%
  summarise(
    avg_pct_aian  = weighted.mean(pct_aian,  w = total_pop, na.rm = TRUE),
    avg_pct_asian = weighted.mean(pct_asian, w = total_pop, na.rm = TRUE),
    avg_pct_nhpi  = weighted.mean(pct_nhpi,  w = total_pop, na.rm = TRUE),
    avg_pct_aapi  = weighted.mean(pct_aapi,  w = total_pop, na.rm = TRUE),
    .groups = "drop"
  )

# Sanity check: percentages must be 0-100; one row per county
cat("Counties in race_county:", nrow(race_county), "\n")
cat("Range avg_pct_aapi:", round(range(race_county$avg_pct_aapi), 2), "\n")
cat("Range avg_pct_aian:", round(range(race_county$avg_pct_aian), 2), "\n")
if (any(race_county$avg_pct_aapi > 100, na.rm = TRUE) || any(race_county$avg_pct_aian > 100, na.rm = TRUE)) {
  stop("A county percentage exceeds 100% — check upstream arithmetic.")
}
if (any(is.na(race_county$avg_pct_aapi)) || any(is.na(race_county$avg_pct_aian))) {
  stop("NA county-level percentage detected — one or more counties may have zero total population after tract filtering. Investigate before proceeding.")
}

# --- Step 2: Join childcare points to county race shares and rural flag ---
rural_lookup <- joined_rural %>%
  select(longitude, latitude, rural) %>%
  distinct()

analysis_data <- childcare_raw %>%
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  left_join(race_county,  by = "county_fips") %>%
  left_join(rural_lookup, by = c("longitude", "latitude")) %>%
  filter(
    !is.na(adj_supply),
    !is.na(avg_pct_asian), !is.na(avg_pct_nhpi),
    !is.na(avg_pct_aapi),  !is.na(avg_pct_aian),
    !is.na(rural),
    state_name != "Puerto Rico"
  ) %>%
  mutate(
    desert = adj_supply <= desert_threshold,
    # Non-exclusive 25% flags — a county can be TRUE on more than one
    is_25_asian  = avg_pct_asian >= 25,
    is_25_nhpi   = avg_pct_nhpi  >= 25,
    is_25_aapi   = avg_pct_aapi  >= 25,
    is_25_aian   = avg_pct_aian  >= 25
  )

# Join quality summary
cat("Rows retained after filtering:", nrow(analysis_data), "\n")
cat("Rows missing race data:",
    sum(is.na(analysis_data$avg_pct_aapi) | is.na(analysis_data$avg_pct_aian)), "\n")
cat("Rows missing rural flag:", sum(is.na(analysis_data$rural)), "\n")
```

## Chart output: share of young children (ages 5 and under) in licensed childcare deserts

```{r chart}
# Compute desert rate (Total, Urban, Rural) for a given non-exclusive flag.
# Each row in analysis_data = 10 synthetic children.
compute_bars <- function(data, flag_col, label) {
  flag_col <- enquo(flag_col)
  subset   <- data %>% filter(!!flag_col)

  total <- subset %>%
    summarise(
      pct_children_in_desert = round(mean(desert) * 100, 1),
      n_points               = n(),
      n_counties             = n_distinct(county_fips)
    ) %>%
    mutate(urbanicity = "Nationwide")

  by_urban <- subset %>%
    mutate(urbanicity = if_else(rural == 1, "Rural", "Urban")) %>%
    group_by(urbanicity) %>%
    summarise(
      pct_children_in_desert = round(mean(desert) * 100, 1),
      n_points               = n(),
      n_counties             = n_distinct(county_fips),
      .groups = "drop"
    )

  bind_rows(total, by_urban) %>%
    mutate(group = label) %>%
    select(group, urbanicity, pct_children_in_desert, n_points, n_counties)
}

chart_data <- bind_rows(
  compute_bars(analysis_data, is_25_aapi,  "AAPI (combined)"),
  compute_bars(analysis_data, is_25_asian, "Asian"),
  compute_bars(analysis_data, is_25_nhpi,  "NHPI"),
  compute_bars(analysis_data, is_25_aian,  "AIAN")
) %>%
  mutate(urbanicity = factor(urbanicity, levels = c("Urban", "Rural", "Nationwide"))) %>%
  arrange(group, urbanicity)

# Sanity check re: Small population counts by county for NPHI alone
if (any(chart_data$pct_children_in_desert > 100 | chart_data$pct_children_in_desert < 0, na.rm=TRUE)) {
  stop("A desert rate is outside 0-100% — check upstream data.")
}

na_rows <- chart_data[is.na(chart_data$pct_children_in_desert), ]
if (nrow(na_rows) > 0) {
  warning(paste(nrow(na_rows), "row(s) have NA for pct_children_in_desert"))
  print(na_rows)
}

chart_data_filtered <- chart_data |>
  filter(!(group == "NHPI" & urbanicity == "Nationwide"))

# Flag text for chart/table footnote
nhpi_flag <- "* Native Hawaiian and Pacific Islander (NHPI) data is not available at the national level due to insufficient sample size."

# Long format (with n_points and n_counties for QA)
print(chart_data_filtered, n = Inf)

# Wide format for Datawrapper / Excel
chart_data_wide <- chart_data_filtered %>%
  select(group, urbanicity, pct_children_in_desert) %>%
  pivot_wider(names_from = urbanicity, values_from = pct_children_in_desert)

print(chart_data_wide)
```

## Diagnostic check: small-county artifacts

BIG FLAG: Groups backed by very few counties can produce misleadingly extreme desert
rates. This flags any group × urbanicity combination where the number of distinct counties is 3 or fewer.

```{r diagnostic-small-n}
chart_data_filtered %>%
  filter(n_counties <= 3) %>%
  select(group, urbanicity, pct_children_in_desert, n_points, n_counties) %>%
  { if (nrow(.) == 0) cat("No small-county flags.\n") else print(.) }
```

