---
title: "Child Care Deserts × Counties"
author: "Hailey Gibbs"
date: "May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Part 1: County-Level Summary

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

### Load Libraries

```{r libraries}
library(tidyverse)
library(tigris)
library(sf)
library(scales)
library(leaflet)
library(leaflet.extras)
library(htmlwidgets)
library(htmltools)
library(maps)
library(here)

# Caching avoids re-downloading files every time you run the script
options(tigris_use_cache = TRUE)
```

### Load and Prepare Child Care Data

```{r load-data}
# Load child care data: CCD full file
childcare_data <- read.csv("cap_ccaccess_2025_revised_02282026.csv", fileEncoding = "UTF-8") %>%
  mutate(county_name = sapply(county_name, function(x) {
    rawToChar(as.raw(utf8ToInt(x)))
  }))

# Set desert parameters
desert_threshold <- 0.33

# Pad FIPS codes to ensure correct string matching
childcare_data <- childcare_data %>%
  mutate(
    is_desert  = adj_supply <= desert_threshold,
    state_fips = str_pad(state_fips,  width = 2, pad = "0"),
    geoid      = str_pad(county_fips, width = 5, pad = "0")
  )
```

### Aggregate to the County Level (Non-CT)

CT is excluded here because Connecticut switched to planning regions in 2022 and requires a separate spatial join workflow below.

```{r county-summary}
county_summary <- childcare_data %>%
  filter(!state_abbrv %in% c("CT", "PR")) %>%
  group_by(geoid, state_fips, state_name, state_abbrv, county_name) %>%
  summarise(
    total_points    = n(),
    desert_points   = sum(is_desert, na.rm = TRUE),
    total_children  = total_points  * 10,
    desert_children = desert_points * 10,
    pct_in_desert   = (desert_children / total_children) * 100,
    .groups = "drop"
  )
```

### Handle Connecticut Planning Regions (Whyyyy did they change these???)

```{r ct-regions}
# Pull CT planning region boundaries from tigris
# counties() returns CT's 9 planning regions (county equivalents since 2022)
# with 5-character GEOIDs that match counties_sf
ct_regions_sf <- counties(
  state = "CT",
  cb    = TRUE,
  year  = 2025
) %>%
  st_transform(crs = 4326) %>%
  select(region_geoid   = GEOID,
         ct_region_name = NAME)

# Isolate CT points and convert to sf
ct_points_sf <- childcare_data %>%
  filter(state_abbrv == "CT") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Spatial join: assign each CT point to its planning region
ct_joined <- st_join(ct_points_sf, ct_regions_sf, join = st_within) %>%
  st_drop_geometry()

# Check for any CT points that didn't fall within a region
ct_unmatched <- ct_joined %>%
  filter(is.na(region_geoid)) %>%
  select(adj_supply, is_desert, county_name, county_fips, geoid)

print(ct_unmatched)
# Warns that 12 CT points did not match a planning region and will be excluded
# They are likely coastal/boundary edge cases
```

```{r ct-snap}
# Recover unmatched CT points using their existing geoid from raw data
ct_snapped <- ct_joined %>%
  filter(is.na(region_geoid)) %>%
  mutate(
    region_geoid   = geoid,
    ct_region_name = county_name
  )

# Combine matched and recovered points
ct_joined_clean <- ct_joined %>%
  filter(!is.na(region_geoid)) %>%
  bind_rows(ct_snapped)

# Verify all 12 are recovered
cat("Total CT points after recovery:", nrow(ct_joined_clean), "\n") # should read 21707
cat("Remaining unmatched:", sum(is.na(ct_joined_clean$region_geoid)), "\n") # should read 0
```

```{r ct-summary}
# Aggregate matched CT points to planning region level
ct_summary <- ct_joined_clean %>%
  group_by(geoid = region_geoid,
           ct_region_name) %>%
  summarise(
    total_points    = n(),
    desert_points   = sum(is_desert, na.rm = TRUE),
    total_children  = total_points  * 10,
    desert_children = desert_points * 10,
    pct_in_desert   = (desert_children / total_children) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    state_fips  = "09",
    state_name  = "Connecticut",
    state_abbrv = "CT",
    county_name = ct_region_name
  ) %>%
  select(-ct_region_name)
```

### Combine County and CT Summaries

```{r full-summary}
full_summary <- bind_rows(county_summary, ct_summary) %>%
  arrange(desc(pct_in_desert))
```

### National Summary Statistics

```{r national-summary}
# National summary statistics (excluding PR)
national_summary <- childcare_data %>%
  filter(state_abbrv != "PR") %>%
  summarise(
    total_children  = n() * 10,
    desert_children = sum(is_desert, na.rm = TRUE) * 10,
    pct_in_desert   = (desert_children / total_children) * 100
  )

print(national_summary)
# Expected output:
#   total_children desert_children pct_in_desert
# 1       22681150        10342090      45.59773
#
# Total children represented:      22,681,150
# Children in child care deserts:  10,342,090
# Percentage in child care desert: 45.6%
```

### County-Level Summary Table

```{r summary-table}
# County-level summary table (includes CT planning regions, excludes PR)
summary_table <- full_summary %>%
  mutate(
    `% in Desert` = round(pct_in_desert, 1),
    `County FIPS`  = substr(geoid, 3, 5)
  ) %>%
  select(
    GEOID                = geoid,
    `County FIPS`,
    State                = state_name,
    `State Abbr`         = state_abbrv,
    County               = county_name,
    `Total Children`     = total_children,
    `Children in Desert` = desert_children,
    `% in Desert`
  )

view(summary_table)
write.csv(summary_table, file = "ccd_by_counties.csv", row.names = FALSE)
```

### Map County Geometries

```{r county-geometries}
# Get county geometries from tigris
counties_sf <- counties(
  cb         = TRUE,
  year       = 2025,
  resolution = "5m"
) %>%
  select(geoid = GEOID, geometry) %>%
  filter(!substr(geoid, 1, 2) %in% c("60", "66", "69", "72", "78")) # Drop territories (incl. PR)

# Join summary data to county geometries
counties_mapped <- counties_sf %>%
  left_join(full_summary, by = "geoid")

# Shift Alaska & Hawaii for contiguous US map
counties_mapped <- shift_geometry(counties_mapped)
```

---

##Interactive Map with Tooltips
### Prepare Data for Map

The map uses `full_summary` from Part 1 directly, which includes the correct CT planning regions and excludes PR. No re-aggregation from raw data is needed.

### Pull County Geometries and Join

```{r map-geometries}
# Pull county geometries
counties_sf <- counties(cb = TRUE, year = 2025, resolution = "5m") %>%
  mutate(geoid = as.character(GEOID)) %>%
  select(geoid, geometry) %>%
  filter(!substr(geoid, 1, 2) %in% c("60", "66", "69", "72", "78")) %>%
  st_transform(crs = 4326)

# Ensures we're not dropping counties from tigris::fips_codes
county_ref <- fips_codes %>%
  mutate(geoid = paste0(state_code, county_code)) %>%
  select(
    geoid,
    county_name_ref = county,
    state_abbrv_ref = state,
    state_name_ref  = state_name
  )

# Join full_summary (includes CT planning regions, excludes PR) to geometries
counties_mapped <- counties_sf %>%
  left_join(full_summary, by = "geoid") %>%
  left_join(county_ref, by = "geoid") %>%
  mutate(
    county_name = coalesce(county_name, county_name_ref),
    state_abbrv = coalesce(state_abbrv, state_abbrv_ref),
    state_name  = coalesce(state_name,  state_name_ref)
  ) %>%
  select(-county_name_ref, -state_abbrv_ref, -state_name_ref)
```

### Define Color Palette

```{r color-palette}
# Define color palette — range derived from full_summary (CT + all non-PR counties)
pct_range <- range(full_summary$pct_in_desert, na.rm = TRUE)

pal <- colorNumeric(
  palette  = c("#2166ac", "#92c5de", "#f7f7f7", "#f4a582", "#d6604d", "#b2182b"),
  domain   = pct_range,
  na.color = "#d0d0d0"
)
```

### Build Tooltips

```{r tooltips}
# Build tooltip
counties_mapped <- counties_mapped %>%
  mutate(
    tooltip_html = ifelse(
      is.na(pct_in_desert),
      paste0(
        "<div style='font-family:\"Segoe UI\", sans-serif; padding:4px 2px;'>",
        "<b style='font-size:13px;'>", county_name, ", ", state_abbrv, "</b><br/>",
        "<span style='font-size:11px; color:#888;'>No data available</span>",
        "</div>"
      ),
      paste0(
        "<div style='font-family:\"Segoe UI\", sans-serif; padding:4px 2px; min-width:200px;'>",

        "<b style='font-size:14px;'>", county_name, ", ", state_abbrv, "</b>",
        "<hr style='margin:5px 0; border:none; border-top:1px solid #ddd;'/>",

        "<div style='margin-bottom:6px;'>",
        "<span style='font-size:11px; color:#555; text-transform:uppercase;",
        "      letter-spacing:0.5px;'>In a Licensed Child Care Desert</span><br/>",
        "<span style='font-size:22px; font-weight:700; color:",
        ifelse(pct_in_desert >= 66, "#b2182b",
               ifelse(pct_in_desert >= 33, "#e08214", "#2166ac")),
        ";'>", sprintf("%.1f%%", pct_in_desert), "</span>",
        "</div>",

        "<table style='width:100%; font-size:12px; border-collapse:collapse;'>",

        "<tr style='background:#f9f9f9;'>",
        "<td style='padding:3px 6px; color:#444;'>Total children represented</td>",
        "<td style='padding:3px 6px; text-align:right; font-weight:600;'>",
        comma(total_children), "</td></tr>",

        "<tr>",
        "<td style='padding:3px 6px; color:#444;'>Children in desert</td>",
        "<td style='padding:3px 6px; text-align:right; font-weight:600;'>",
        comma(desert_children), "</td></tr>",

        "</table></div>"
      )
    )
  )
```

### National Statistics (for External Use)

Not displayed on the map — use for external tables or reports.

```{r national-stats}
national <- full_summary %>%
  summarise(
    total  = sum(total_children,  na.rm = TRUE),
    desert = sum(desert_children, na.rm = TRUE),
    pct    = desert / total * 100
  )
```

### Map Title HTML

```{r title-html}
title_html <- tagList(
  tags$div(
    style = paste(
      "position:absolute; top:10px; left:50px; z-index:1000;",
      "background:white; padding:10px 16px; border-radius:6px;",
      "box-shadow:0 2px 8px rgba(0,0,0,0.2);",
      "font-family:'Segoe UI', sans-serif; max-width:340px;"
    ),
    tags$b(style = "font-size:13px;", "Child Care Desert Map"),
    tags$br(),
    if (!is.na(national$pct)) {
      tags$span(
        style = "font-size:11px; color:#555;",
        sprintf(
          "Nationally, %.1f%% of young children (%s of %s) live in a child care desert (adj_supply ≤ %.2f)",
          national$pct,
          comma(national$desert),
          comma(national$total),
          desert_threshold
        )
      )
    }
  )
)
```

### Build and Display Leaflet Map

```{r leaflet-map}
map <- leaflet(
  counties_mapped,
  options = leafletOptions(minZoom = 3, maxZoom = 12)
) %>%

  addProviderTiles("CartoDB.Positron",    group = "Light (default)") %>%
  addProviderTiles("CartoDB.DarkMatter",  group = "Dark") %>%
  addProviderTiles("Esri.WorldStreetMap", group = "Street") %>%

  setView(lng = -96, lat = 38, zoom = 4) %>%

  addPolygons(
    fillColor   = ~pal(pct_in_desert),
    fillOpacity = 0.75,
    color       = "white",
    weight      = 0.4,
    opacity     = 0.8,
    highlightOptions = highlightOptions(
      weight       = 2,
      color        = "#333",
      fillOpacity  = 0.95,
      bringToFront = TRUE
    ),
    label        = ~lapply(tooltip_html, HTML),
    labelOptions = labelOptions(
      style = list(
        "box-shadow"    = "3px 3px 6px rgba(0,0,0,0.25)",
        "border-radius" = "5px",
        "padding"       = "6px 10px"
      ),
      textsize  = "13px",
      direction = "auto"
    ),
    group = "Counties"
  ) %>%

  addLegend(
    position  = "bottomright",
    pal       = pal,
    values    = pct_range,
    title     = HTML("% of Children<br/>"),
    labFormat = labelFormat(suffix = "%", digits = 0),
    opacity   = 0.95
  ) %>%

  addLayersControl(
    baseGroups    = c("Light (default)", "Dark", "Street"),
    overlayGroups = "Counties",
    options       = layersControlOptions(collapsed = TRUE)
  ) %>%

  addSearchOSM(
    options = searchOptions(
      collapsed            = TRUE,
      autoCollapse         = TRUE,
      zoom                 = 8,
      hideMarkerOnCollapse = TRUE
    )
  ) %>%

  addResetMapButton() %>%
  addFullscreenControl()
  # addControl(title_html, position = "topleft")

# Display map in viewer pane
map
```

### Export Map as Self-Contained HTML

```{r save-widget, eval=FALSE}
# here::here() resolves to the project root regardless of where knit() runs,
# ensuring the file is always written to the expected location.
saveWidget(
  widget        = map,
  file          = here::here("childcare_desert_interactive_map.html"),
  selfcontained = TRUE,
  title         = "Child Care Desert Map by County"
)
```
