---
title: "Child Care Desert Estimates by School District"
author: "Hailey Gibbs"
date: "May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Setup

```{r libraries-and-data}
library(tidyverse)

desert_threshold      <- 0.33
low_sample_n_points    <- 5   # districts with fewer than this many lat/long points
                               # (i.e. < 50 synthetic children) are flagged as low_sample.
                               # Adjust this threshold if you want a stricter/looser cutoff.

childcare_data <- read_csv("cap_ccaccess_2025_revised_02282026.csv")
```

---

## National / Overall Estimates

```{r national-desert-rate}
national_desert_rate <- childcare_data %>%
  filter(!is.na(adj_supply)) %>%
  mutate(
    desert = adj_supply <= desert_threshold
  ) %>%
  summarise(
    total_children  = n() * 10,
    desert_children = sum(desert) * 10,
    pct_desert      = round(mean(desert) * 100, 1)
  )

print(national_desert_rate)
```

---

## State Estimates

### Desert Rate by State

```{r desert-by-state}
desert_by_state <- childcare_data %>%
  filter(!is.na(adj_supply)) %>%                       # Fixed: exclude NA adj_supply so the
                                                         # denominator matches the numerator
  mutate(in_desert = adj_supply <= desert_threshold) %>%
  group_by(state_name) %>%
  summarise(
    total_children      = n() * 10,                    # Fixed: was n() (missing *10 for synthetic children)
    children_in_desert  = sum(in_desert) * 10,          # Fixed: was sum(in_desert, na.rm = TRUE) with no *10
    percent_in_desert   = round((children_in_desert / total_children) * 100, 1),
    mean_adj_supply     = mean(adj_supply),             # na.rm no longer needed - NAs already filtered above
    .groups = "drop"
  ) %>%
  arrange(desc(percent_in_desert))

view(desert_by_state)
```

### Desert and Low Supply Comparison by State

```{r desert-comparison}
desert_comparison <- childcare_data %>%
  filter(!is.na(adj_supply)) %>%                        # Fixed: exclude NA adj_supply so the
                                                          # denominator matches the numerator
  group_by(state_name) %>%
  summarise(
    total_children  = n() * 10,                          # Fixed: was n() (missing *10 for synthetic children)
    pct_zero_supply = (sum(adj_supply == 0)   / n()) * 100,
    pct_low_supply  = (sum(adj_supply <  0.1) / n()) * 100,
    pct_desert      = (sum(adj_supply <= desert_threshold) / n()) * 100,
    mean_supply     = mean(adj_supply),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_desert))

view(desert_comparison)
write_csv(desert_comparison, "deserts_comparison.csv")
```

---

## Deserts by Care Type

```{r desert-by-care-type}
national_desert_rates <- childcare_data %>%
  mutate(
    desert_hs   = adj_supply_hs   <= desert_threshold,
    desert_prek = adj_supply_prek <= desert_threshold,
    desert_fcc  = adj_supply_fcc  <= desert_threshold,
    desert_ccc  = adj_supply_ccc  <= desert_threshold,
    desert_any  = adj_supply      <= desert_threshold
  ) %>%
  summarise(
    total_children      = n() * 10,
    `Overall`           = round(mean(desert_any,  na.rm = TRUE) * 100, 1),
    `Head Start`        = round(mean(desert_hs,   na.rm = TRUE) * 100, 1),
    `Pre-K`             = round(mean(desert_prek, na.rm = TRUE) * 100, 1),
    `Family Child Care` = round(mean(desert_fcc,  na.rm = TRUE) * 100, 1),
    `Child Care Center` = round(mean(desert_ccc,  na.rm = TRUE) * 100, 1)
  )

print(national_desert_rates)
```

---

## School District Estimates

Districts are identified by `state_name` + `schooldistrict_code` + `sd_name` together, since
district codes are only guaranteed unique within a state. Both tables below include an
`n_points` column (the number of lat/long points, i.e. groups of 10 synthetic children,
underlying each district) and a `low_sample` flag marking districts with fewer than
`low_sample_n_points` points, so small-sample districts are visible rather than silently
mixed in with everything else. Filter or treat `low_sample == TRUE` rows however makes sense
for your use case (e.g. exclude from a public-facing map, footnote them in a table, etc.).

### Desert Rate by School District

```{r desert-by-district}
desert_by_district <- childcare_data %>%
  filter(!is.na(adj_supply)) %>%
  mutate(in_desert = adj_supply <= desert_threshold) %>%
  group_by(state_name, schooldistrict_code, sd_name) %>%
  summarise(
    n_points            = n(),
    total_children      = n() * 10,
    children_in_desert  = sum(in_desert) * 10,
    percent_in_desert   = round((children_in_desert / total_children) * 100, 1),
    mean_adj_supply     = mean(adj_supply),
    .groups = "drop"
  ) %>%
  mutate(low_sample = n_points < low_sample_n_points) %>%
  arrange(desc(percent_in_desert))

view(desert_by_district)
write_csv(desert_by_district, "deserts_by_district.csv")
```

### Deserts by Care Type, by School District

```{r desert-by-district-caretype}
desert_by_district_caretype <- childcare_data %>%
  group_by(state_name, schooldistrict_code, sd_name) %>%
  summarise(
    n_points            = n(),
    total_children      = n() * 10,
    `Overall`           = round(mean(adj_supply      <= desert_threshold, na.rm = TRUE) * 100, 1),
    `Head Start`        = round(mean(adj_supply_hs   <= desert_threshold, na.rm = TRUE) * 100, 1),
    `Pre-K`             = round(mean(adj_supply_prek <= desert_threshold, na.rm = TRUE) * 100, 1),
    `Family Child Care` = round(mean(adj_supply_fcc  <= desert_threshold, na.rm = TRUE) * 100, 1),
    `Child Care Center` = round(mean(adj_supply_ccc  <= desert_threshold, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  mutate(low_sample = n_points < low_sample_n_points) %>%
  arrange(desc(Overall))

view(desert_by_district_caretype)
write_csv(desert_by_district_caretype, "deserts_by_district_caretype.csv")
```

### National and State Roll-Ups from District Estimates (Population-Weighted)

The `percent_in_desert` and care-type columns in the two tables above are already
population-correct *within* each district, since every row of `childcare_data` represents
an equal 10 children. The distortion shows up one level higher, when districts themselves
are rolled up into a state or national number: a plain `mean()` across districts gives a
1-point district with 10 children the same weight as a district with 500,000 children, so
the many small/rural districts (which skew toward high desert rates) inflate the average.
The chunks below roll `desert_by_district` and `desert_by_district_caretype` up to the
state and national level two ways — weighted by `total_children` (correct) and as a plain,
unweighted average across districts (shown only so you can see the size of the distortion)
— so you don't have to go back to the raw `childcare_data` to get a trustworthy top-line
number from the district-level files.

```{r district-rollup-national}
national_from_districts <- desert_by_district %>%
  summarise(
    n_districts            = n(),
    total_children         = sum(total_children),
    children_in_desert     = sum(children_in_desert),
    pct_desert_weighted    = round((children_in_desert / total_children) * 100, 1),  # correct
    pct_desert_unweighted  = round(mean(percent_in_desert), 1)                       # naive - for comparison only
  )

print(national_from_districts)
```

```{r district-rollup-state}
state_from_districts <- desert_by_district %>%
  group_by(state_name) %>%
  summarise(
    n_districts            = n(),
    total_children         = sum(total_children),
    children_in_desert     = sum(children_in_desert),
    pct_desert_weighted    = round((children_in_desert / total_children) * 100, 1),  # correct
    pct_desert_unweighted  = round(mean(percent_in_desert), 1),                      # naive - for comparison only
    .groups = "drop"
  ) %>%
  arrange(desc(pct_desert_weighted))

# pct_desert_weighted here should match percent_in_desert in desert_by_state closely -
# useful as a consistency check between the two pipelines.
view(state_from_districts)
write_csv(state_from_districts, "deserts_by_state_from_districts.csv")
```

```{r district-rollup-caretype-national}
caretype_national_from_districts <- desert_by_district_caretype %>%
  summarise(
    n_districts       = n(),
    total_children    = sum(total_children),
    `Overall`           = round(sum(Overall           / 100 * total_children) / sum(total_children) * 100, 1),
    `Head Start`        = round(sum(`Head Start`       / 100 * total_children) / sum(total_children) * 100, 1),
    `Pre-K`             = round(sum(`Pre-K`            / 100 * total_children) / sum(total_children) * 100, 1),
    `Family Child Care` = round(sum(`Family Child Care`/ 100 * total_children) / sum(total_children) * 100, 1),
    `Child Care Center` = round(sum(`Child Care Center`/ 100 * total_children) / sum(total_children) * 100, 1),
    `Overall (unweighted avg)`           = round(mean(Overall), 1),
    `Head Start (unweighted avg)`        = round(mean(`Head Start`), 1),
    `Pre-K (unweighted avg)`             = round(mean(`Pre-K`), 1),
    `Family Child Care (unweighted avg)` = round(mean(`Family Child Care`), 1),
    `Child Care Center (unweighted avg)` = round(mean(`Child Care Center`), 1)
  )

print(caretype_national_from_districts)
```
