---
title: "Head Start Deserts and No-Supply Areas by State and Rurality"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)

desert_threshold <- 0.33
```

## 1. Load data

```{r load-data}
poverty_data   <- read_csv("poverty_total_and_under5_clean.csv", show_col_types = FALSE)
childcare_data <- read_csv("cap_ccaccess_2025_revised_02282026.csv", show_col_types = FALSE)
joined_rural   <- read_csv("joined_rural.csv", show_col_types = FALSE)
```

## 2. Clean poverty data (tract → county)

```{r clean-poverty}
poverty_county <- poverty_data %>%
  mutate(county_fips = str_sub(GEOID, 1, 5)) %>%
  filter(!is.na(pct_poverty_under5)) %>%
  group_by(county_fips) %>%
  summarise(
    avg_pct_poverty_under5 = mean(pct_poverty_under5, na.rm = TRUE),
    .groups = "drop"
  )
```

## 3. Join childcare data with poverty and rurality

```{r join-data}
childcare_with_poverty <- childcare_data %>%
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  left_join(poverty_county, by = "county_fips")

rural_lookup <- joined_rural %>%
  select(longitude, latitude, rural) %>%
  distinct()

childcare_with_poverty <- childcare_with_poverty %>%
  left_join(rural_lookup, by = c("longitude", "latitude"))

cat("Rows with rural status:", sum(!is.na(childcare_with_poverty$rural)), "\n")
cat("Rows missing rural status:", sum(is.na(childcare_with_poverty$rural)), "\n")
```

## 4. Drop missing values and compute flags

Each row represents 10 synthetic children. `eligible_children` scales that by
the county's poverty rate under 5. `hs_desert` flags Head Start deserts
(adjusted supply at or below the desert threshold). `hs_no_access` flags areas
with literally zero Head Start supply.

```{r compute-flags}
childcare_with_poverty <- childcare_with_poverty %>%
  filter(!is.na(adj_supply_hs), !is.na(avg_pct_poverty_under5), !is.na(rural)) %>%
  mutate(
    eligible_children = 10 * (avg_pct_poverty_under5 / 100),
    hs_desert    = adj_supply_hs <= desert_threshold,
    hs_no_access = adj_supply_hs == 0
  )
```

## 5. Restrict to the 50 states + DC

`childcare_data` already includes clean `state_name` / `state_abbrv` columns,
so we don't need to join a crosswalk (doing so would collide with the
existing columns and rename them). Instead we use the crosswalk only as a
lookup of valid state FIPS codes to filter out territories.

```{r state-crosswalk}
state_fips_crosswalk <- tribble(
  ~state_fips, ~state_name,            ~state_abbr,
  "01", "Alabama",               "AL",
  "02", "Alaska",                "AK",
  "04", "Arizona",               "AZ",
  "05", "Arkansas",              "AR",
  "06", "California",            "CA",
  "08", "Colorado",              "CO",
  "09", "Connecticut",           "CT",
  "10", "Delaware",              "DE",
  "11", "District of Columbia",  "DC",
  "12", "Florida",               "FL",
  "13", "Georgia",               "GA",
  "15", "Hawaii",                "HI",
  "16", "Idaho",                 "ID",
  "17", "Illinois",              "IL",
  "18", "Indiana",               "IN",
  "19", "Iowa",                  "IA",
  "20", "Kansas",                "KS",
  "21", "Kentucky",              "KY",
  "22", "Louisiana",             "LA",
  "23", "Maine",                 "ME",
  "24", "Maryland",              "MD",
  "25", "Massachusetts",         "MA",
  "26", "Michigan",              "MI",
  "27", "Minnesota",             "MN",
  "28", "Mississippi",           "MS",
  "29", "Missouri",              "MO",
  "30", "Montana",               "MT",
  "31", "Nebraska",              "NE",
  "32", "Nevada",                "NV",
  "33", "New Hampshire",         "NH",
  "34", "New Jersey",            "NJ",
  "35", "New Mexico",            "NM",
  "36", "New York",              "NY",
  "37", "North Carolina",        "NC",
  "38", "North Dakota",          "ND",
  "39", "Ohio",                  "OH",
  "40", "Oklahoma",              "OK",
  "41", "Oregon",                "OR",
  "42", "Pennsylvania",          "PA",
  "44", "Rhode Island",          "RI",
  "45", "South Carolina",        "SC",
  "46", "South Dakota",          "SD",
  "47", "Tennessee",             "TN",
  "48", "Texas",                 "TX",
  "49", "Utah",                  "UT",
  "50", "Vermont",               "VT",
  "51", "Virginia",              "VA",
  "53", "Washington",            "WA",
  "54", "West Virginia",         "WV",
  "55", "Wisconsin",             "WI",
  "56", "Wyoming",               "WY"
)

childcare_with_poverty_state <- childcare_with_poverty %>%
  mutate(state_fips = str_sub(county_fips, 1, 2)) %>%
  filter(state_fips %in% state_fips_crosswalk$state_fips) %>%
  mutate(hs_no_access = adj_supply_hs == 0)

# Sanity check: state_name should have survived intact (no join collision)
stopifnot("state_name" %in% names(childcare_with_poverty_state))

cat(
  "Rows before state restriction:", nrow(childcare_with_poverty), "\n",
  "Rows after restricting to 50 states + DC:", nrow(childcare_with_poverty_state), "\n",
  "Rows dropped (territories / unmatched FIPS):",
  nrow(childcare_with_poverty) - nrow(childcare_with_poverty_state), "\n"
)
```

## 6. Table 1 — National summary, by rurality

Share of poverty-eligible children (under 5) living in a Head Start desert,
and share living in an area with zero Head Start supply, broken out by
urban/rural status.

```{r table-national-rurality}
hs_national_by_rurality <- childcare_with_poverty_state %>%
  mutate(urbanicity = if_else(rural == 1, "Rural", "Urban")) %>%
  group_by(urbanicity) %>%
  summarise(
    total_eligible_children     = sum(eligible_children),
    desert_eligible_children    = sum(eligible_children[hs_desert]),
    no_access_eligible_children = sum(eligible_children[hs_no_access]),
    pct_eligible_in_desert      = round((desert_eligible_children / total_eligible_children) * 100, 1),
    pct_eligible_no_access      = round((no_access_eligible_children / total_eligible_children) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(urbanicity)

kable(hs_national_by_rurality,
      caption = "Head Start deserts and no-supply areas, nationwide, by rurality")
```

## 7. Table 2 — By state (alphabetical)

```{r table-by-state}
hs_eligible_desert_by_state <- childcare_with_poverty_state %>%
  group_by(state_name, state_abbrv) %>%
  summarise(
    total_eligible_children      = sum(eligible_children),
    desert_eligible_children     = sum(eligible_children[hs_desert]),
    no_access_eligible_children  = sum(eligible_children[hs_no_access]),
    pct_eligible_in_desert       = round((desert_eligible_children / total_eligible_children) * 100, 1),
    pct_eligible_no_access       = round((no_access_eligible_children / total_eligible_children) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(state_name)

kable(hs_eligible_desert_by_state,
      caption = "Head Start deserts and no-supply areas, by state (alphabetical)")
```

## 8. Table 3 — By state and rurality

```{r table-by-state-rurality}
hs_eligible_desert_by_state_rurality <- childcare_with_poverty_state %>%
  mutate(urbanicity = if_else(rural == 1, "Rural", "Urban")) %>%
  group_by(state_name, state_abbrv, urbanicity) %>%
  summarise(
    total_eligible_children      = sum(eligible_children),
    desert_eligible_children     = sum(eligible_children[hs_desert]),
    no_access_eligible_children  = sum(eligible_children[hs_no_access]),
    pct_eligible_in_desert       = round((desert_eligible_children / total_eligible_children) * 100, 1),
    pct_eligible_no_access       = round((no_access_eligible_children / total_eligible_children) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(state_name, urbanicity)

kable(hs_eligible_desert_by_state_rurality,
      caption = "Head Start deserts and no-supply areas, by state and rurality")
```

## 9. Export tables to CSV

```{r export-csv}
write_csv(hs_national_by_rurality, "hs_national_by_rurality.csv")
write_csv(hs_eligible_desert_by_state, "hs_by_state.csv")
write_csv(hs_eligible_desert_by_state_rurality, "hs_by_state_rurality.csv")
```
