---
title: "Head Start Access Where Other Licensed Care Is Entirely Absent"
subtitle: "Poverty-Eligible Children, by State and Rurality"
author: "Hailey Gibbs"
date: "September 16, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(scales)
library(readr)

children_per_row <- 10 # based on ACS 5-year block group data
```

## 1. Load data

```{r load-data}
poverty_data   <- read_csv("poverty_total_and_under5_clean.csv",
                            col_types = cols(GEOID = col_character()))
childcare_data <- read_csv("cap_ccaccess_2025_revised_02282026.csv")
joined_rural   <- read_csv("joined_rural.csv")
```

## 2. Fix truncated GEOID codes, then build county-level poverty rate

## Flagging issue in previous analyses:
`GEOID` was stored as a number upstream at some point, which strips the
leading zero for the seven states whose FIPS code is a single digit (01 AL,
02 AK, 04 AZ, 05 AR, 06 CA, 08 CO, 09 CT). Left uncorrected, this shifts
`str_sub(GEOID, 1, 5)` by one character for those states' tracts, producing a
`county_fips` that cannot match anything downstream -- which silently dropped
all seven states out of the analytic sample. Confirmed via `nchar(GEOID)`:
tract-level GEOIDs should be a uniform 11 characters (2 state + 3 county + 6
tract); the affected rows were 10. Padding restores the correct value.

```{r fix-and-clean-poverty}
poverty_data <- poverty_data %>%
  mutate(GEOID = str_pad(GEOID, width = 11, pad = "0"))

geoid_len <- table(nchar(poverty_data$GEOID))
if (length(geoid_len) > 1) {
  warning("poverty_data$GEOID is not a uniform 11 characters after padding -- ",
          "county identifiers derived from it may still be misaligned for ",
          "some rows. Length counts: ",
          paste(names(geoid_len), geoid_len, sep = "=", collapse = ", "))
}

candidate_weight_cols <- c("total_under5", "under5_total", "total_pop_under5",
                            "pop_under5", "under5_population", "total_under5_pop")
weight_col <- intersect(candidate_weight_cols, names(poverty_data))

if (length(weight_col) >= 1) {
  weight_col <- weight_col[1]
  cat("Using '", weight_col, "' as the tract-level under-5 population weight.\n", sep = "")
} else {
  weight_col <- NA
  warning(
    "No recognized under-5 population column found in poverty_data. Falling ",
    "back to an unweighted mean of pct_poverty_under5 across tracts. Run ",
    "colnames(poverty_data) and add the correct column name to ",
    "candidate_weight_cols if one exists under a different name."
  )
}

poverty_tract <- poverty_data %>%
  mutate(county_fips = str_sub(GEOID, 1, 5)) %>%
  filter(!is.na(pct_poverty_under5))

if (!is.na(weight_col)) {
  poverty_county <- poverty_tract %>%
    filter(!is.na(.data[[weight_col]])) %>%
    group_by(county_fips) %>%
    summarise(
      avg_pct_poverty_under5 = weighted.mean(pct_poverty_under5, w = .data[[weight_col]], na.rm = TRUE),
      .groups = "drop"
    )
} else {
  poverty_county <- poverty_tract %>%
    group_by(county_fips) %>%
    summarise(avg_pct_poverty_under5 = mean(pct_poverty_under5, na.rm = TRUE), .groups = "drop")
}
```

## 3. Merge in rurality data

```{r merge-rurality}
normalize_lon <- function(lon) ((lon + 180) %% 360) - 180

childcare_data_norm <- childcare_data %>%
  mutate(
    lon_key = round(normalize_lon(longitude), 5),
    lat_key = round(latitude, 5)
  )

rural_lookup <- joined_rural %>%
  select(longitude, latitude, rural) %>%
  mutate(
    lon_key = round(normalize_lon(longitude), 5),
    lat_key = round(latitude, 5)
  ) %>%
  select(lon_key, lat_key, rural) %>%
  distinct()

dup_coords <- rural_lookup %>% count(lon_key, lat_key) %>% filter(n > 1)
if (nrow(dup_coords) > 0) {
  rural_lookup <- rural_lookup %>% distinct(lon_key, lat_key, .keep_all = TRUE)
  warning(nrow(dup_coords), " coordinate pairs had conflicting rural codes after ",
          "normalization -- kept the first value for each.")
}

childcare_with_rural <- childcare_data_norm %>%
  left_join(rural_lookup, by = c("lon_key", "lat_key"))

cat("Rows before join:", nrow(childcare_data), " | Rows after join:",
    nrow(childcare_with_rural), "\n")
cat("Rows missing rural status:", sum(is.na(childcare_with_rural$rural)), "\n")
```

## 4. Merge in poverty data

```{r merge-poverty}
childcare_with_poverty <- childcare_with_rural %>%
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  left_join(poverty_county, by = "county_fips")

cat("Rows with a matched county poverty rate:   ",
    sum(!is.na(childcare_with_poverty$avg_pct_poverty_under5)), "\n")
cat("Rows missing a matched county poverty rate:",
    sum(is.na(childcare_with_poverty$avg_pct_poverty_under5)), "\n")
```

## 5. Build the analytic sample

```{r build-analytic-sample}
n_start <- nrow(childcare_with_poverty)

step1_no_pr <- childcare_with_poverty %>%
  filter(!(as.character(state_fips) %in% c("72", "072")))

step2_complete_supply <- step1_no_pr %>%
  filter(
    !is.na(adj_supply_hs), !is.na(adj_supply_fcc),
    !is.na(adj_supply_ccc), !is.na(adj_supply_prek)
  )

step3_has_poverty <- step2_complete_supply %>%
  filter(!is.na(avg_pct_poverty_under5))

childcare_clean <- step3_has_poverty %>%
  filter(!is.na(rural))

cat("Starting rows:                           ", n_start, "\n")
cat("Dropped -- Puerto Rico:                   ", n_start - nrow(step1_no_pr), "\n")
cat("Dropped -- missing adj_supply_* value:    ", nrow(step1_no_pr) - nrow(step2_complete_supply), "\n")
cat("Dropped -- no matched county poverty rate:", nrow(step2_complete_supply) - nrow(step3_has_poverty), "\n")
cat("Dropped -- no matched rurality status:    ", nrow(step3_has_poverty) - nrow(childcare_clean), "\n")
cat("Remaining (analytic sample):              ", nrow(childcare_clean), "\n")

# Guard against a repeat of the GEOID bug (or any other silent full-state
# dropout): flag any state present upstream that ends up with zero rows in
# the final analytic sample, since a fully-dropped state produces no NA to
# find -- it just disappears from every table below without a trace.
dropped_states <- childcare_with_poverty %>%
  distinct(state_fips, state_name) %>%
  anti_join(childcare_clean %>% distinct(state_fips), by = "state_fips")

if (nrow(dropped_states) > 0) {
  warning("The following states have zero rows in the analytic sample and ",
          "will not appear in any table below -- confirm this is expected ",
          "(e.g. a territory) and not a repeat join failure: ",
          paste(dropped_states$state_name, collapse = ", "))
} else {
  cat("All", n_distinct(childcare_with_poverty$state_fips),
      "states/territories present upstream are represented in the analytic sample.\n")
}
```

## 6. Eligibility, desert flag, and Head Start access flag

```{r flag-and-weight}
childcare_clean <- childcare_clean %>%
  mutate(
    eligible_children  = children_per_row * (avg_pct_poverty_under5 / 100),
    other_all_zero     = adj_supply_fcc == 0 & adj_supply_ccc == 0 & adj_supply_prek == 0,
    hs_any             = adj_supply_hs > 0,
    hs_any_others_zero = hs_any & other_all_zero
  )
```

## 7. National summary

```{r summarize-national}
national_summary <- childcare_clean %>%
  summarise(
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert              = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible        = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2)
  )

cat("National summary:\n")
national_summary %>% as_tibble() %>% print(n = Inf)
```

## 8. Summary by rurality (national)

```{r summarize-by-rural}
rural_summary <- childcare_clean %>%
  group_by(rural) %>%
  summarise(
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert       = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = if_else(
      eligible_in_other_zero > 0,
      round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2),
      NA_real_
    ),
    .groups = "drop"
  )

cat("Summary by rurality:\n")
rural_summary %>% as_tibble() %>% print(n = Inf)
```

## 9. Summary by state (eligible children)

```{r summarize-by-state}
state_summary <- childcare_clean %>%
  group_by(state_fips) %>%
  summarise(
    state_name    = first(state_name),
    state_abbrv   = first(state_abbrv),
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert       = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = if_else(
      eligible_in_other_zero > 0,
      round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2),
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_of_other_zero_eligible))

state_summary %>%
  transmute(
    State = state_name,
    Abbrv = state_abbrv,
    `Eligible children (total)`                    = comma(round(total_eligible_children)),
    `Eligible children in desert (FCC=CCC=PreK=0)` = comma(round(eligible_in_other_zero)),
    `% of eligible children in desert`             = pct_in_desert,
    `Eligible children in desert + any HS`         = comma(round(eligible_in_hs_any_others_zero)),
    `% of desert-eligible with any HS`             = pct_of_other_zero_eligible
  ) %>%
  as_tibble() %>%
  print(n = Inf)
```

## 10. Chart: percentage by state

```{r chart-by-state, fig.width=8, fig.height=10}
state_summary %>%
  filter(!is.na(pct_of_other_zero_eligible)) %>%
  ggplot(aes(x = fct_reorder(state_abbrv, pct_of_other_zero_eligible),
             y = pct_of_other_zero_eligible)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Head Start Presence Where All Other Licensed Care Is Absent",
    subtitle = "% of poverty-eligible zero-others children with any Head Start supply, by state",
    x = "State",
    y = "% of zero-others eligible children with any Head Start supply"
  ) +
  theme_minimal(base_size = 11)
```

## 11. Summary by state and rurality (eligible children)

```{r summarize-by-state-rural}
state_rural_summary <- childcare_clean %>%
  group_by(state_fips, rural) %>%
  summarise(
    state_name  = first(state_name),
    state_abbrv = first(state_abbrv),
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert       = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = if_else(
      eligible_in_other_zero > 0,
      round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2),
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(state_fips, rural)

state_rural_summary %>%
  transmute(
    State = state_name,
    Abbrv = state_abbrv,
    Rural = rural,
    `Eligible children (total)`                    = comma(round(total_eligible_children)),
    `Eligible children in desert (FCC=CCC=PreK=0)` = comma(round(eligible_in_other_zero)),
    `% of eligible children in desert`             = pct_in_desert,
    `Eligible children in desert + any HS`         = comma(round(eligible_in_hs_any_others_zero)),
    `% of desert-eligible with any HS`             = pct_of_other_zero_eligible,
    `% of all eligible with any HS-only access`    = pct_of_all_eligible
  ) %>%
  as_tibble() %>%
  print(n = Inf)
```

## 12. Chart: percentage by state, split by rurality

```{r chart-by-state-rural, fig.width=9, fig.height=11}
state_rural_summary %>%
  filter(!is.na(pct_of_other_zero_eligible)) %>%
  ggplot(aes(x = fct_reorder(state_abbrv, pct_of_other_zero_eligible, .fun = max),
             y = pct_of_other_zero_eligible, fill = rural)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Head Start Presence Where All Other Licensed Care Is Absent",
    subtitle = "By state and rurality",
    x = "State",
    y = "% of zero-others eligible children with any Head Start supply",
    fill = "Rural"
  ) +
  theme_minimal(base_size = 11)
```

## 13. Summary interpretation

```{r summary-stats, include=FALSE}
n_states <- nrow(state_summary)
ranked   <- state_summary %>% filter(!is.na(pct_of_other_zero_eligible))
top_state    <- ranked %>% slice_max(pct_of_other_zero_eligible, n = 1)
bottom_state <- ranked %>% slice_min(pct_of_other_zero_eligible, n = 1)

# Pulled dynamically rather than hard-coded, so this works regardless of how
# the `rural` column happens to be encoded (logical, character, factor, etc.)
desert_high <- rural_summary %>% slice_max(pct_in_desert, n = 1)
desert_low  <- rural_summary %>% slice_min(pct_in_desert, n = 1)
desert_ratio <- round(desert_high$pct_in_desert / desert_low$pct_in_desert, 1)

# Ranks states by pct_of_all_eligible (share of ALL poverty-eligible children
# with Head Start-only access) within the rural = 1 group specifically -- this
# is the metric the "roughly 10 times more likely" / state-list claims below
# are based on, not pct_of_other_zero_eligible (which is printed in section 11
# but describes a different denominator: share of *desert-dwelling* children).
rural_state_rank <- state_rural_summary %>%
  filter(rural == 1) %>%
  filter(!is.na(pct_of_all_eligible)) %>%
  arrange(desc(pct_of_all_eligible))

top5_rural_states    <- rural_state_rank %>% slice_max(pct_of_all_eligible, n = 5)
bottom5_rural_states <- rural_state_rank %>% slice_min(pct_of_all_eligible, n = 5)

national_rural_ratio <- round(
  (rural_summary %>% filter(rural == desert_high$rural) %>% pull(pct_of_all_eligible)) /
  (rural_summary %>% filter(rural == desert_low$rural)  %>% pull(pct_of_all_eligible)),
  1
)

# Largest rural-vs-urban gap in pct_of_all_eligible, among states with both a
# rural and a non-rural row and a non-zero non-rural rate (so the ratio is defined).
gap_by_state <- state_rural_summary %>%
  filter(!is.na(pct_of_all_eligible)) %>%
  select(state_name, state_abbrv, rural, pct_of_all_eligible) %>%
  pivot_wider(names_from = rural, values_from = pct_of_all_eligible, names_prefix = "rural_") %>%
  filter(!is.na(rural_0), !is.na(rural_1), rural_0 > 0) %>%
  mutate(gap_ratio = round(rural_1 / rural_0, 1)) %>%
  arrange(desc(gap_ratio))

largest_gap_state <- gap_by_state %>% slice_max(gap_ratio, n = 1)
```

## 14. Export summary tables

```{r export, eval=FALSE}
write_csv(state_summary, "headstart_any_access_zero_others_poverty_weighted_by_state_v5.csv")
write_csv(rural_summary, "headstart_any_access_zero_others_poverty_weighted_by_rurality_v5.csv")
write_csv(state_rural_summary, "headstart_any_access_zero_others_poverty_weighted_by_state_and_rurality_v5.csv")
```

## 15. Figure 1: Per 1,000 income-eligible children, by rurality

Figure 1 is built from `rural_summary` (section 8), so it updates whenever the
data changes. Each panel is a 100 x 10 grid of 1,000 squares. The shares are rounded **once, at
the desert level**, and the Head Start–only count is rounded separately.
"In a desert, no Head Start" is the difference between those two rounded numbers.
That way the squares always add up to 1,000, and they always match the labels.

```{r fig1-per-1000, fig.width=10, fig.height=5.6}
# ---- Settings to edit ---------------------------------------------------------
term_label  <- "deserts with no access to licensed home-based care, center-based care, or preschool"
save_png    <- TRUE       # also write a high-resolution PNG for layout
grid_cols   <- 100        # 100 x 10 = 1,000 squares per panel

# ---- Map the rural codes to labels (and stop if the coding is unexpected) ----
rural_codes <- as.character(rural_summary$rural)
if (!all(rural_codes %in% c("0", "1", "FALSE", "TRUE"))) {
  stop("Unexpected values in rural_summary$rural: ",
       paste(unique(rural_codes), collapse = ", "),
       ". Update the label mapping in the fig1-per-1000 chunk.")
}

fig1_rates <- rural_summary %>%
  mutate(
    group = if_else(as.character(rural) %in% c("1", "TRUE"), "Rural", "Nonrural"),
    group = factor(group, levels = c("Rural", "Nonrural")),
    # Unrounded rates per 1,000 income-eligible children
    desert_per_1000  = 1000 * eligible_in_other_zero         / total_eligible_children,
    hs_only_per_1000 = 1000 * eligible_in_hs_any_others_zero / total_eligible_children,
    # Round once, then take the difference, so squares and labels always agree
    desert_n  = round(desert_per_1000),
    hs_only_n = round(hs_only_per_1000),
    desert_no_hs_n = desert_n - hs_only_n,
    other_care_n   = 1000 - desert_n
  )

# Print the unrounded and rounded numbers so the figure can be checked by hand
fig1_rates %>%
  select(group, desert_per_1000, hs_only_per_1000,
         desert_n, hs_only_n, desert_no_hs_n, other_care_n) %>%
  mutate(across(where(is.double), ~ round(.x, 1))) %>%
  print()

stopifnot(all(fig1_rates$hs_only_n + fig1_rates$desert_no_hs_n +
              fig1_rates$other_care_n == 1000))

# ---- Build the 1,000-square grid for each panel -------------------------------
cat_levels <- c(
  "Head Start is the only licensed option",
  str_wrap(paste0("No Head Start, in ", term_label), 50),
  "Other licensed care nearby"
)

fig1_grid <- fig1_rates %>%
  select(group, hs_only_n, desert_no_hs_n, other_care_n) %>%
  pivot_longer(-group, names_to = "cat", values_to = "n") %>%
  mutate(cat = factor(recode(cat,
                             hs_only_n      = cat_levels[1],
                             desert_no_hs_n = cat_levels[2],
                             other_care_n   = cat_levels[3]),
                      levels = cat_levels)) %>%
  arrange(group, cat) %>%
  uncount(n) %>%
  group_by(group) %>%
  mutate(idx = row_number() - 1,
         x   = idx %% grid_cols,     # fill left to right...
         y   = idx %/% grid_cols) %>%  # ...then bottom to top
  ungroup()

# Panel titles carry the rounded numbers (direct labels, so the chart doesn't rely on color alone)
panel_labels <- fig1_rates %>%
  transmute(group,
            lab = paste0(group, ": about ", comma(desert_n), " live in ", term_label,
                         ";\nfor about ", comma(hs_only_n),
                         ", Head Start is the only licensed option"))
panel_labeller <- as_labeller(setNames(panel_labels$lab, as.character(panel_labels$group)))

# One hue, light to dark: the darker the square, the fewer licensed options nearby
fig1_colors <- setNames(c("#0B4F6C", "#6FA8C4", "#E3E6EA"), cat_levels)

fig1 <- ggplot(fig1_grid, aes(x = x, y = y, fill = cat)) +
  geom_tile(width = 0.82, height = 0.82) +
  facet_wrap(~ group, ncol = 1, labeller = panel_labeller) +
  scale_fill_manual(values = fig1_colors, name = NULL) +
  coord_equal(expand = FALSE) +
  labs(
    title    = "Rural children are far more likely to live where Head Start is the only licensed option",
    subtitle = "Estimated outcomes per 1,000 income-eligible children under age 5",
    caption  = paste0(
      "Note: Estimates, rounded to the nearest child per 1,000. Licensed child care deserts are areas with at least three young children\n",
      "per licensed slot. Access reflects the proximity of licensed supply, not enrollment. ",
      "Nonrural includes urban and suburban areas.\n",
      "Source: Center for American Progress analysis of licensed child care deserts data (2025) and ",
      "American Community Survey 2019–2023 five-year estimates."
    )
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13, margin = margin(b = 4)),
    plot.subtitle    = element_text(color = "grey30", margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey35", size = 8, hjust = 0,
                                    margin = margin(t = 10), lineheight = 1.1),
    plot.caption.position = "plot",
    plot.title.position   = "plot",
    strip.text       = element_text(face = "bold", hjust = 0, size = 10.5,
                                    margin = margin(t = 8, b = 4)),
    legend.position  = "top",
    legend.justification = "left",
    legend.key.size  = unit(10, "pt"),
    plot.background  = element_rect(fill = "white", color = NA),
    plot.margin      = margin(12, 12, 12, 12)
  )

fig1

if (save_png) {
  ggsave("fig1_per_1000_by_rurality.png", fig1, width = 10, height = 5.6, dpi = 300, bg = "white")
}
```

## 16. Figure 2: Rural share of each group of income-eligible children

Figure 2 is built from `state_rural_summary` (section 11), so it updates
whenever the data changes. Each bar splits one group of income-eligible
children into rural and nonrural. The rural share is rounded to a whole
percent, and the nonrural share is 100 minus that, so each bar always totals
100.

To see whether the Alaska and Hawaii values move the figure, I'd need to add them to
`exclude_states` (for example, `c("AK", "HI")`). The chunk prints the shares
with and without the excluded states

```{r fig2-rural-share, fig.width=10, fig.height=4.8}
# ---- Settings to edit ---------------------------------------------------------
term_label     <- "deserts with no access to licensed home-based care, center-based care, or preschool"
exclude_states <- c()        # e.g., c("AK", "HI") pending the join check
save_png       <- TRUE       # also write a high-resolution PNG for layout

# ---- Map the rural codes to labels (and stop if the coding is unexpected) ----
rural_codes <- as.character(state_rural_summary$rural)
if (!all(rural_codes %in% c("0", "1", "FALSE", "TRUE"))) {
  stop("Unexpected values in state_rural_summary$rural: ",
       paste(unique(rural_codes), collapse = ", "),
       ". Update the label mapping in the fig2-rural-share chunk.")
}

# ---- Rural share of each group ------------------------------------------------
rural_shares <- function(df) {
  df %>%
    mutate(is_rural = as.character(rural) %in% c("1", "TRUE")) %>%
    summarise(
      all_eligible = sum(total_eligible_children[is_rural])        / sum(total_eligible_children),
      no_other     = sum(eligible_in_other_zero[is_rural])         / sum(eligible_in_other_zero),
      hs_only      = sum(eligible_in_hs_any_others_zero[is_rural]) / sum(eligible_in_hs_any_others_zero)
    ) %>%
    pivot_longer(everything(), names_to = "measure", values_to = "rural_share")
}

shares_all  <- rural_shares(state_rural_summary)
shares_used <- rural_shares(state_rural_summary %>% filter(!state_abbrv %in% exclude_states))

# Print unrounded shares (in percent) so the figure can be checked by hand
shares_all %>%
  rename(rural_pct_all_states = rural_share) %>%
  left_join(shares_used %>% rename(rural_pct_in_figure = rural_share), by = "measure") %>%
  mutate(across(where(is.double), ~ round(100 * .x, 1))) %>%
  print()
if (length(exclude_states) > 0) {
  cat("Figure excludes:", paste(exclude_states, collapse = ", "), "\n")
}

# ---- Build the bars ----------------------------------------------------------
measure_labels <- c(
  all_eligible = "All income-eligible children",
  no_other     = str_wrap(paste0("Children in ", term_label), 38),
  hs_only      = "Children for whom Head Start is the\nonly licensed option"
)

fig2_data <- shares_used %>%
  mutate(
    rural_pct    = round(100 * rural_share),
    nonrural_pct = 100 - rural_pct,          # each bar totals exactly 100
    measure      = factor(measure_labels[measure], levels = rev(measure_labels))
  ) %>%
  select(measure, Rural = rural_pct, Nonrural = nonrural_pct) %>%
  pivot_longer(c(Rural, Nonrural), names_to = "group", values_to = "pct") %>%
  mutate(group = factor(group, levels = c("Nonrural", "Rural"))) %>%  # rural sits at the left
  group_by(measure) %>%
  arrange(desc(group), .by_group = TRUE) %>%
  mutate(xmid = cumsum(pct) - pct / 2) %>%
  ungroup()

stopifnot(all(tapply(fig2_data$pct, fig2_data$measure, sum) == 100))

fig2_colors <- c(Rural = "#0B4F6C", Nonrural = "#D5DADF")

fig2 <- ggplot(fig2_data, aes(x = pct, y = measure, fill = group)) +
  geom_col(width = 0.62, color = "white", linewidth = 0.8) +
  geom_text(aes(x = xmid, label = paste0(pct, "%"),
                color = group),
            size = 4, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(values = fig2_colors, breaks = c("Rural", "Nonrural"), name = NULL) +
  scale_color_manual(values = c(Rural = "white", Nonrural = "grey20")) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 100.5)) +
  labs(
    title    = "Rural children are 19 percent of income-eligible children but most of\nthose with no other licensed child care option",
    subtitle = "Rural and nonrural shares of income-eligible children under age 5, by group",
    x = NULL, y = NULL,
    caption  = paste0(
      "Note: Estimates, rounded to the nearest percent. Licensed child care deserts are areas with at least three young children per licensed slot.\n",
      "Access reflects the proximity of licensed supply, not enrollment. Nonrural includes urban and suburban areas.",
      if (length(exclude_states) > 0) paste0(" Excludes ", paste(exclude_states, collapse = ", "), ".") else "",
      "\nSource: Center for American Progress analysis of licensed child care deserts data (2025) and ",
      "American Community Survey 2019–2023 five-year estimates."
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 12.5, margin = margin(b = 4)),
    plot.subtitle      = element_text(color = "grey30", margin = margin(b = 8)),
    plot.caption       = element_text(color = "grey35", size = 8, hjust = 0,
                                      margin = margin(t = 10), lineheight = 1.1),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    axis.text.x        = element_blank(),
    axis.text.y        = element_text(size = 10.5, color = "grey15", hjust = 1),
    panel.grid         = element_blank(),
    legend.position    = "top",
    legend.justification = "left",
    legend.key.size    = unit(10, "pt"),
    plot.background    = element_rect(fill = "white", color = NA),
    plot.margin        = margin(12, 16, 12, 12)
  )

fig2

if (save_png) {
  ggsave("fig2_rural_share_by_group.png", fig2, width = 10, height = 4.8, dpi = 300, bg = "white")
}
```
