---
title: "Head Start Access Where Other Licensed Care Is Entirely Absent"
subtitle: "Poverty-Eligible Children, by State and Rurality"
author: "Hailey Gibbs"
date: "September 16, 2026 (revised October 1, 2026)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(scales)
library(readr)
library(tidycensus)
library(sf)
library(tigris)

options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)   # planar point-in-polygon is fine at tract scale and much faster

children_per_row <- 10 # based on ACS 5-year block group data

# One-time setup, if not already done:
# tidycensus::census_api_key("efe9c4d9098021681847e000144aa4b99487cd9c", install = TRUE); readRenviron("~/.Renviron")
```

## Revision note: how poverty is now assigned

The earlier version of this analysis had two problems with the poverty rates
used to estimate income-eligible children.

1. **Smoothing within counties.** Tract-level `pct_poverty_under5` was averaged
   up to the county and joined to every childcare row by `county_fips`, so every
   row in a county got the same rate. In counties that mix rural and urban
   areas, this erased the within-county differences that the rural/nonrural
   comparison depends on.
2. **No population weights.** The poverty file contains only percentages
   (`GEOID`, `NAME`, `pct_poverty_total`, `pct_poverty_under5`, `flag`), so the
   county "average" was an unweighted mean of tract percentages: a tract with
   50 young children counted as much as one with 2,000.

This version fixes both:

- Poverty **counts** come from ACS table B17001 (2019–2023 five-year), for
  children under 5 below poverty and all children under 5 whose poverty status
  is determined. Any rate built from summed counts is correctly weighted.
- Each childcare row is placed in its **census tract** using its coordinates,
  and gets that tract's rate. The county rate (also built from summed counts)
  is used only as a labeled fallback when a tract has no usable data.

## 1. Load data

```{r load-data}
childcare_data <- read_csv("cap_ccaccess_2025_revised_02282026.csv")
joined_rural   <- read_csv("joined_rural.csv")

# The original percent-only file is kept for a validation check in Section 2.
old_poverty_file <- "poverty_total_and_under5_clean.csv"
```

## 2. Tract and county poverty counts (ACS B17001)

B17001 is poverty status in the past 12 months by sex by age.

- `_004` / `_018`: male / female under 5, income below poverty
- `_033` / `_047`: male / female under 5, income at or above poverty

Tract geography must be requested one state at a time, so the download takes
a few minutes. The result is cached to an `.rds` file so later knits skip it.

```{r poverty-counts}
# ---- Settings to edit ---------------------------------------------------------
acs_year         <- 2023  # 2019-2023 five-year estimates
min_tract_under5 <- 1     # tracts with fewer children under 5 than this use the
                          # county rate instead (raise, e.g. to 30, to avoid
                          # very noisy small-tract estimates)
tract_counts_cache <- paste0("acs_b17001_under5_tracts_", acs_year, ".rds")

# 50 states + DC (Puerto Rico is excluded from the analysis)
state_codes <- unique(tidycensus::fips_codes$state_code)
state_codes <- state_codes[as.integer(state_codes) <= 56]

b17001_vars <- c(
  pov_m    = "B17001_004",
  pov_f    = "B17001_018",
  nonpov_m = "B17001_033",
  nonpov_f = "B17001_047"
)

if (file.exists(tract_counts_cache)) {
  poverty_counts_tract <- readRDS(tract_counts_cache)
} else {
  poverty_counts_tract <- map_dfr(state_codes, ~ get_acs(
    geography = "tract",
    variables = b17001_vars,
    state     = .x,
    year      = acs_year,
    survey    = "acs5",
    output    = "wide"
  )) %>%
    transmute(
      GEOID,  # 11-character string with leading zeros intact
      under5_in_poverty = pov_mE + pov_fE,
      under5_total      = pov_mE + pov_fE + nonpov_mE + nonpov_fE,
      pct_poverty_tract = if_else(under5_total > 0,
                                  100 * under5_in_poverty / under5_total,
                                  NA_real_)
    )
  saveRDS(poverty_counts_tract, tract_counts_cache)
}

stopifnot(all(nchar(poverty_counts_tract$GEOID) == 11))

# County rates from summed counts -- weighted by construction
poverty_county <- poverty_counts_tract %>%
  mutate(county_fips = str_sub(GEOID, 1, 5)) %>%
  group_by(county_fips) %>%
  summarise(
    county_under5_in_poverty = sum(under5_in_poverty),
    county_under5_total      = sum(under5_total),
    .groups = "drop"
  ) %>%
  mutate(pct_poverty_county = if_else(county_under5_total > 0,
                                      100 * county_under5_in_poverty / county_under5_total,
                                      NA_real_))

cat("Tracts:", nrow(poverty_counts_tract),
    " | Tracts with zero children under 5:", sum(poverty_counts_tract$under5_total == 0),
    " | Counties:", nrow(poverty_county), "\n")
```

### Validation against the original percent file

If the original `pct_poverty_under5` came from the same table and vintage,
nearly all tracts should match, with rounding-sized differences.

```{r validate-old-poverty}
if (file.exists(old_poverty_file)) {
  read_csv(old_poverty_file, col_types = cols(GEOID = col_character())) %>%
    mutate(GEOID = str_pad(GEOID, width = 11, pad = "0")) %>%
    inner_join(poverty_counts_tract, by = "GEOID") %>%
    mutate(diff = abs(pct_poverty_under5 - pct_poverty_tract)) %>%
    summarise(
      tracts_matched     = n(),
      share_within_0.5pt = round(mean(diff < 0.5, na.rm = TRUE), 3),
      max_abs_diff       = round(max(diff, na.rm = TRUE), 2)
    ) %>%
    print()
} else {
  cat("Original poverty file not found; skipping validation.\n")
}
```

## 3. Merge in rurality data

```{r merge-rurality}
normalize_lon <- function(lon) ((lon + 180) %% 360) - 180

childcare_data_norm <- childcare_data %>%
  mutate(
    lon_key = round(normalize_lon(longitude), 5),
    lat_key = round(latitude, 5)
  )

rural_lookup <- joined_rural %>%
  select(longitude, latitude, rural) %>%
  mutate(
    lon_key = round(normalize_lon(longitude), 5),
    lat_key = round(latitude, 5)
  ) %>%
  select(lon_key, lat_key, rural) %>%
  distinct()

dup_coords <- rural_lookup %>% count(lon_key, lat_key) %>% filter(n > 1)
if (nrow(dup_coords) > 0) {
  rural_lookup <- rural_lookup %>% distinct(lon_key, lat_key, .keep_all = TRUE)
  warning(nrow(dup_coords), " coordinate pairs had conflicting rural codes after ",
          "normalization -- kept the first value for each.")
}

childcare_with_rural <- childcare_data_norm %>%
  left_join(rural_lookup, by = c("lon_key", "lat_key"))

cat("Rows before join:", nrow(childcare_data), " | Rows after join:",
    nrow(childcare_with_rural), "\n")
cat("Rows missing rural status:", sum(is.na(childcare_with_rural$rural)), "\n")
```

## 4. Assign each row its tract poverty rate

Each row's coordinates are placed inside a 2020 census tract (the geography of
the 2019–2023 ACS), and the row gets that tract's rate. Rows that fall outside
every tract polygon, or in a tract below `min_tract_under5`, fall back to the
county rate; `poverty_source` records which was used.

```{r tract-boundaries}
tract_shapes_cache <- paste0("tract_shapes_", acs_year, ".rds")

# cb = TRUE is the smaller cartographic file, clipped to the shoreline. If many
# coastal rows come back unmatched below, switch to cb = FALSE and delete the cache.
if (file.exists(tract_shapes_cache)) {
  tracts_sf <- readRDS(tract_shapes_cache)
} else {
  tracts_sf <- map(state_codes, ~ tracts(state = .x, year = acs_year, cb = TRUE)) %>%
    do.call(rbind, .) %>%
    select(tract_geoid = GEOID)
  saveRDS(tracts_sf, tract_shapes_cache)
}
```

```{r merge-poverty}
childcare_pts <- childcare_with_rural %>%
  mutate(row_id = row_number()) %>%
  st_as_sf(coords = c("lon_key", "lat_key"), crs = 4326,
           remove = FALSE, na.fail = FALSE) %>%
  st_transform(st_crs(tracts_sf))

childcare_tract <- childcare_pts %>%
  st_join(tracts_sf, join = st_within, left = TRUE) %>%
  st_drop_geometry() %>%
  distinct(row_id, .keep_all = TRUE)  # guard against points on a shared boundary

stopifnot(nrow(childcare_tract) == nrow(childcare_with_rural))

childcare_with_poverty <- childcare_tract %>%
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  left_join(poverty_counts_tract %>%
              select(tract_geoid = GEOID,
                     tract_under5_total = under5_total,
                     pct_poverty_tract),
            by = "tract_geoid") %>%
  left_join(poverty_county %>% select(county_fips, pct_poverty_county),
            by = "county_fips") %>%
  mutate(
    use_tract = !is.na(pct_poverty_tract) & tract_under5_total >= min_tract_under5,
    poverty_source = case_when(
      use_tract                  ~ "tract",
      !is.na(pct_poverty_county) ~ "county fallback",
      TRUE                       ~ "none"
    ),
    pct_poverty_under5 = if_else(use_tract, pct_poverty_tract, pct_poverty_county)
  )

cat("Rows not inside any tract polygon:", sum(is.na(childcare_with_poverty$tract_geoid)), "\n")
childcare_with_poverty %>% count(poverty_source) %>% print()

# If fallback rows are concentrated in one rurality group, they could bias the
# rural/nonrural comparison and are worth a closer look.
childcare_with_poverty %>% count(rural, poverty_source) %>% print()
```

### Connecticut county codes

From 2022 on, the ACS uses Connecticut's nine planning regions (09110–09190) as
county equivalents instead of the eight old counties (09001–09015). The tract
join is unaffected because it matches by coordinates, but the county fallback
only works if the childcare file uses the same codes.

```{r ct-check}
ct_childcare <- sort(unique(childcare_with_poverty$county_fips[
  str_starts(childcare_with_poverty$county_fips, "09")]))
ct_acs <- sort(poverty_county$county_fips[str_starts(poverty_county$county_fips, "09")])

cat("CT county codes in childcare file:", paste(ct_childcare, collapse = ", "), "\n")
cat("CT county codes in ACS counts:    ", paste(ct_acs, collapse = ", "), "\n")
if (length(setdiff(ct_childcare, ct_acs)) > 0) {
  warning("Connecticut rows use county codes not in the ACS; those rows can use ",
          "tract rates but have no county fallback.")
}
```

## 5. Build the analytic sample

```{r build-analytic-sample}
n_start <- nrow(childcare_with_poverty)

step1_no_pr <- childcare_with_poverty %>%
  filter(!(as.character(state_fips) %in% c("72", "072")))

step2_complete_supply <- step1_no_pr %>%
  filter(
    !is.na(adj_supply_hs), !is.na(adj_supply_fcc),
    !is.na(adj_supply_ccc), !is.na(adj_supply_prek)
  )

step3_has_poverty <- step2_complete_supply %>%
  filter(!is.na(pct_poverty_under5))

childcare_clean <- step3_has_poverty %>%
  filter(!is.na(rural))

cat("Starting rows:                           ", n_start, "\n")
cat("Dropped -- Puerto Rico:                   ", n_start - nrow(step1_no_pr), "\n")
cat("Dropped -- missing adj_supply_* value:    ", nrow(step1_no_pr) - nrow(step2_complete_supply), "\n")
cat("Dropped -- no tract or county poverty rate:", nrow(step2_complete_supply) - nrow(step3_has_poverty), "\n")
cat("Dropped -- no matched rurality status:    ", nrow(step3_has_poverty) - nrow(childcare_clean), "\n")
cat("Remaining (analytic sample):              ", nrow(childcare_clean), "\n")
cat("Analytic rows using tract rate:           ", sum(childcare_clean$poverty_source == "tract"), "\n")
cat("Analytic rows using county fallback:      ", sum(childcare_clean$poverty_source == "county fallback"), "\n")

# Flag any state present upstream that ends up with zero rows in the final
# analytic sample, since a fully-dropped state produces no NA to find -- it
# just disappears from every table below without a trace.
dropped_states <- childcare_with_poverty %>%
  distinct(state_fips, state_name) %>%
  anti_join(childcare_clean %>% distinct(state_fips), by = "state_fips")

if (nrow(dropped_states) > 0) {
  warning("The following states have zero rows in the analytic sample and ",
          "will not appear in any table below -- confirm this is expected ",
          "(e.g. a territory) and not a join failure: ",
          paste(dropped_states$state_name, collapse = ", "))
} else {
  cat("All", n_distinct(childcare_with_poverty$state_fips),
      "states/territories present upstream are represented in the analytic sample.\n")
}
```

## 6. Eligibility, desert flag, and Head Start access flag

`eligible_children` uses each row's tract rate (or county fallback).
`eligible_children_county_only` uses the county rate for every row; it exists
only for the sensitivity check in Section 8b.

```{r flag-and-weight}
childcare_clean <- childcare_clean %>%
  mutate(
    eligible_children             = children_per_row * (pct_poverty_under5 / 100),
    eligible_children_county_only = children_per_row * (pct_poverty_county / 100),
    other_all_zero     = adj_supply_fcc == 0 & adj_supply_ccc == 0 & adj_supply_prek == 0,
    hs_any             = adj_supply_hs > 0,
    hs_any_others_zero = hs_any & other_all_zero
  )
```

## 7. National summary

```{r summarize-national}
national_summary <- childcare_clean %>%
  summarise(
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert              = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible        = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2)
  )

cat("National summary:\n")
national_summary %>% as_tibble() %>% print(n = Inf)
```

## 8. Summary by rurality (national)

```{r summarize-by-rural}
rural_summary <- childcare_clean %>%
  group_by(rural) %>%
  summarise(
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert       = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = if_else(
      eligible_in_other_zero > 0,
      round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2),
      NA_real_
    ),
    .groups = "drop"
  )

cat("Summary by rurality:\n")
rural_summary %>% as_tibble() %>% print(n = Inf)
```

## 8b. Sensitivity: tract rates vs. county-only rates

How much did county smoothing matter? This compares the rural share of each
group, and the rural/nonrural desert rates, under tract-level rates versus a
single (weighted) county rate for every row. Rows without a county rate are
left out of the county-only column.

```{r sensitivity-tract-vs-county}
compare_weights <- function(df, w) {
  df %>%
    filter(!is.na({{ w }})) %>%
    mutate(is_rural = as.character(rural) %in% c("1", "TRUE"), wt = {{ w }}) %>%
    summarise(
      rural_share_all_eligible = 100 * sum(wt[is_rural]) / sum(wt),
      rural_share_desert       = 100 * sum(wt[is_rural & other_all_zero]) / sum(wt[other_all_zero]),
      rural_share_hs_only      = 100 * sum(wt[is_rural & hs_any_others_zero]) / sum(wt[hs_any_others_zero]),
      rural_desert_per_1000    = 1000 * sum(wt[is_rural & other_all_zero])  / sum(wt[is_rural]),
      nonrural_desert_per_1000 = 1000 * sum(wt[!is_rural & other_all_zero]) / sum(wt[!is_rural]),
      rural_hs_only_per_1000   = 1000 * sum(wt[is_rural & hs_any_others_zero])  / sum(wt[is_rural]),
      nonrural_hs_only_per_1000 = 1000 * sum(wt[!is_rural & hs_any_others_zero]) / sum(wt[!is_rural])
    ) %>%
    pivot_longer(everything(), names_to = "measure")
}

compare_weights(childcare_clean, eligible_children) %>%
  rename(tract_rates = value) %>%
  left_join(compare_weights(childcare_clean, eligible_children_county_only) %>%
              rename(county_only_rates = value),
            by = "measure") %>%
  mutate(difference = tract_rates - county_only_rates,
         across(where(is.double), ~ round(.x, 1))) %>%
  print()
```

## 9. Summary by state (eligible children)

```{r summarize-by-state}
state_summary <- childcare_clean %>%
  group_by(state_fips) %>%
  summarise(
    state_name    = first(state_name),
    state_abbrv   = first(state_abbrv),
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert       = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = if_else(
      eligible_in_other_zero > 0,
      round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2),
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_of_other_zero_eligible))

state_summary %>%
  transmute(
    State = state_name,
    Abbrv = state_abbrv,
    `Eligible children (total)`                    = comma(round(total_eligible_children)),
    `Eligible children in desert (FCC=CCC=PreK=0)` = comma(round(eligible_in_other_zero)),
    `% of eligible children in desert`             = pct_in_desert,
    `Eligible children in desert + any HS`         = comma(round(eligible_in_hs_any_others_zero)),
    `% of desert-eligible with any HS`             = pct_of_other_zero_eligible
  ) %>%
  as_tibble() %>%
  print(n = Inf)
```

## 10. Chart: percentage by state

```{r chart-by-state, fig.width=8, fig.height=10}
state_summary %>%
  filter(!is.na(pct_of_other_zero_eligible)) %>%
  ggplot(aes(x = fct_reorder(state_abbrv, pct_of_other_zero_eligible),
             y = pct_of_other_zero_eligible)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Head Start Presence Where All Other Licensed Care Is Absent",
    subtitle = "% of poverty-eligible zero-others children with any Head Start supply, by state",
    x = "State",
    y = "% of zero-others eligible children with any Head Start supply"
  ) +
  theme_minimal(base_size = 11)
```

## 11. Summary by state and rurality (eligible children)

```{r summarize-by-state-rural}
state_rural_summary <- childcare_clean %>%
  group_by(state_fips, rural) %>%
  summarise(
    state_name  = first(state_name),
    state_abbrv = first(state_abbrv),
    total_eligible_children        = sum(eligible_children),
    eligible_in_other_zero         = sum(eligible_children[other_all_zero]),
    eligible_in_hs_any_others_zero = sum(eligible_children[hs_any_others_zero]),
    pct_in_desert       = round(100 * eligible_in_other_zero / total_eligible_children, 2),
    pct_of_all_eligible = round(100 * eligible_in_hs_any_others_zero / total_eligible_children, 4),
    pct_of_other_zero_eligible = if_else(
      eligible_in_other_zero > 0,
      round(100 * eligible_in_hs_any_others_zero / eligible_in_other_zero, 2),
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(state_fips, rural)

state_rural_summary %>%
  transmute(
    State = state_name,
    Abbrv = state_abbrv,
    Rural = rural,
    `Eligible children (total)`                    = comma(round(total_eligible_children)),
    `Eligible children in desert (FCC=CCC=PreK=0)` = comma(round(eligible_in_other_zero)),
    `% of eligible children in desert`             = pct_in_desert,
    `Eligible children in desert + any HS`         = comma(round(eligible_in_hs_any_others_zero)),
    `% of desert-eligible with any HS`             = pct_of_other_zero_eligible,
    `% of all eligible with any HS-only access`    = pct_of_all_eligible
  ) %>%
  as_tibble() %>%
  print(n = Inf)
```

## 12. Chart: percentage by state, split by rurality

```{r chart-by-state-rural, fig.width=9, fig.height=11}
state_rural_summary %>%
  filter(!is.na(pct_of_other_zero_eligible)) %>%
  ggplot(aes(x = fct_reorder(state_abbrv, pct_of_other_zero_eligible, .fun = max),
             y = pct_of_other_zero_eligible, fill = rural)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Head Start Presence Where All Other Licensed Care Is Absent",
    subtitle = "By state and rurality",
    x = "State",
    y = "% of zero-others eligible children with any Head Start supply",
    fill = "Rural"
  ) +
  theme_minimal(base_size = 11)
```

## 13. Summary interpretation

```{r summary-stats, include=FALSE}
n_states <- nrow(state_summary)
ranked   <- state_summary %>% filter(!is.na(pct_of_other_zero_eligible))
top_state    <- ranked %>% slice_max(pct_of_other_zero_eligible, n = 1)
bottom_state <- ranked %>% slice_min(pct_of_other_zero_eligible, n = 1)

# Pulled dynamically rather than hard-coded, so this works regardless of how
# the `rural` column happens to be encoded (logical, character, factor, etc.)
desert_high <- rural_summary %>% slice_max(pct_in_desert, n = 1)
desert_low  <- rural_summary %>% slice_min(pct_in_desert, n = 1)
desert_ratio <- round(desert_high$pct_in_desert / desert_low$pct_in_desert, 1)

# Ranks states by pct_of_all_eligible (share of ALL poverty-eligible children
# with Head Start-only access) within the rural = 1 group specifically -- this
# is the metric the "roughly 10 times more likely" / state-list claims below
# are based on, not pct_of_other_zero_eligible (which is printed in section 11
# but describes a different denominator: share of *desert-dwelling* children).
rural_state_rank <- state_rural_summary %>%
  filter(rural == 1) %>%
  filter(!is.na(pct_of_all_eligible)) %>%
  arrange(desc(pct_of_all_eligible))

top5_rural_states    <- rural_state_rank %>% slice_max(pct_of_all_eligible, n = 5)
bottom5_rural_states <- rural_state_rank %>% slice_min(pct_of_all_eligible, n = 5)

national_rural_ratio <- round(
  (rural_summary %>% filter(rural == desert_high$rural) %>% pull(pct_of_all_eligible)) /
  (rural_summary %>% filter(rural == desert_low$rural)  %>% pull(pct_of_all_eligible)),
  1
)

# Largest rural-vs-urban gap in pct_of_all_eligible, among states with both a
# rural and a non-rural row and a non-zero non-rural rate (so the ratio is defined).
gap_by_state <- state_rural_summary %>%
  filter(!is.na(pct_of_all_eligible)) %>%
  select(state_name, state_abbrv, rural, pct_of_all_eligible) %>%
  pivot_wider(names_from = rural, values_from = pct_of_all_eligible, names_prefix = "rural_") %>%
  filter(!is.na(rural_0), !is.na(rural_1), rural_0 > 0) %>%
  mutate(gap_ratio = round(rural_1 / rural_0, 1)) %>%
  arrange(desc(gap_ratio))

largest_gap_state <- gap_by_state %>% slice_max(gap_ratio, n = 1)
```

## 14. Export summary tables

```{r export, eval=FALSE}
write_csv(state_summary, "headstart_any_access_zero_others_tract_poverty_by_state_v6.csv")
write_csv(rural_summary, "headstart_any_access_zero_others_tract_poverty_by_rurality_v6.csv")
write_csv(state_rural_summary, "headstart_any_access_zero_others_tract_poverty_by_state_and_rurality_v6.csv")
```

## 15. Figure 1: Per 1,000 income-eligible children, by rurality

Figure 1 is built from `rural_summary` (section 8), so it updates whenever the
data changes. Each panel is a 100 x 10 grid of 1,000 squares. The shares are rounded **once, at
the desert level**, and the Head Start–only count is rounded separately.
"In a desert, no Head Start" is the difference between those two rounded numbers.
That way the squares always add up to 1,000, and they always match the labels.

```{r fig1-per-1000, fig.width=10, fig.height=5.6}
# ---- Settings to edit ---------------------------------------------------------
term_label  <- "deserts with no access to licensed home-based care, center-based care, or preschool"
save_png    <- TRUE       # also write a high-resolution PNG for layout
grid_cols   <- 100        # 100 x 10 = 1,000 squares per panel

# ---- Map the rural codes to labels (and stop if the coding is unexpected) ----
rural_codes <- as.character(rural_summary$rural)
if (!all(rural_codes %in% c("0", "1", "FALSE", "TRUE"))) {
  stop("Unexpected values in rural_summary$rural: ",
       paste(unique(rural_codes), collapse = ", "),
       ". Update the label mapping in the fig1-per-1000 chunk.")
}

fig1_rates <- rural_summary %>%
  mutate(
    group = if_else(as.character(rural) %in% c("1", "TRUE"), "Rural", "Nonrural"),
    group = factor(group, levels = c("Rural", "Nonrural")),
    # Unrounded rates per 1,000 income-eligible children
    desert_per_1000  = 1000 * eligible_in_other_zero         / total_eligible_children,
    hs_only_per_1000 = 1000 * eligible_in_hs_any_others_zero / total_eligible_children,
    # Round once, then take the difference, so squares and labels always agree
    desert_n  = round(desert_per_1000),
    hs_only_n = round(hs_only_per_1000),
    desert_no_hs_n = desert_n - hs_only_n,
    other_care_n   = 1000 - desert_n
  )

# Print the unrounded and rounded numbers so the figure can be checked by hand
fig1_rates %>%
  select(group, desert_per_1000, hs_only_per_1000,
         desert_n, hs_only_n, desert_no_hs_n, other_care_n) %>%
  mutate(across(where(is.double), ~ round(.x, 1))) %>%
  print()

stopifnot(all(fig1_rates$hs_only_n + fig1_rates$desert_no_hs_n +
              fig1_rates$other_care_n == 1000))

# ---- Build the 1,000-square grid for each panel -------------------------------
cat_levels <- c(
  "Head Start is the only licensed option",
  str_wrap(paste0("No Head Start, in ", term_label), 50),
  "Other licensed care nearby"
)

fig1_grid <- fig1_rates %>%
  select(group, hs_only_n, desert_no_hs_n, other_care_n) %>%
  pivot_longer(-group, names_to = "cat", values_to = "n") %>%
  mutate(cat = factor(recode(cat,
                             hs_only_n      = cat_levels[1],
                             desert_no_hs_n = cat_levels[2],
                             other_care_n   = cat_levels[3]),
                      levels = cat_levels)) %>%
  arrange(group, cat) %>%
  uncount(n) %>%
  group_by(group) %>%
  mutate(idx = row_number() - 1,
         x   = idx %% grid_cols,     # fill left to right...
         y   = idx %/% grid_cols) %>%  # ...then bottom to top
  ungroup()

# Panel titles carry the rounded numbers (direct labels, so the chart doesn't rely on color alone)
panel_labels <- fig1_rates %>%
  transmute(group,
            lab = paste0(group, ": about ", comma(desert_n), " live in ", term_label,
                         ";\nfor about ", comma(hs_only_n),
                         ", Head Start is the only licensed option"))
panel_labeller <- as_labeller(setNames(panel_labels$lab, as.character(panel_labels$group)))

# One hue, light to dark: the darker the square, the fewer licensed options nearby
fig1_colors <- setNames(c("#0B4F6C", "#6FA8C4", "#E3E6EA"), cat_levels)

fig1 <- ggplot(fig1_grid, aes(x = x, y = y, fill = cat)) +
  geom_tile(width = 0.82, height = 0.82) +
  facet_wrap(~ group, ncol = 1, labeller = panel_labeller) +
  scale_fill_manual(values = fig1_colors, name = NULL) +
  coord_equal(expand = FALSE) +
  labs(
    title    = "Rural children are far more likely to live where Head Start is the only licensed option",
    subtitle = "Estimated outcomes per 1,000 income-eligible children under age 5",
    caption  = paste0(
      "Note: Estimates, rounded to the nearest child per 1,000. Licensed child care deserts are areas with at least three young children\n",
      "per licensed slot. Access reflects the proximity of licensed supply, not enrollment. ",
      "Nonrural includes urban and suburban areas.\n",
      "Income eligibility uses census tract poverty rates for children under 5.\n",
      "Source: Center for American Progress analysis of licensed child care deserts data (2025) and ",
      "American Community Survey 2019–2023 five-year estimates."
    )
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13, margin = margin(b = 4)),
    plot.subtitle    = element_text(color = "grey30", margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey35", size = 8, hjust = 0,
                                    margin = margin(t = 10), lineheight = 1.1),
    plot.caption.position = "plot",
    plot.title.position   = "plot",
    strip.text       = element_text(face = "bold", hjust = 0, size = 10.5,
                                    margin = margin(t = 8, b = 4)),
    legend.position  = "top",
    legend.justification = "left",
    legend.key.size  = unit(10, "pt"),
    plot.background  = element_rect(fill = "white", color = NA),
    plot.margin      = margin(12, 12, 12, 12)
  )

fig1

if (save_png) {
  ggsave("fig1_per_1000_by_rurality.png", fig1, width = 10, height = 5.6, dpi = 300, bg = "white")
}
```

## 16. Figure 2: Rural share of each group of income-eligible children

Figure 2 is built from `state_rural_summary` (section 11), so it updates
whenever the data changes. Each bar splits one group of income-eligible
children into rural and nonrural. The rural share is rounded to a whole
percent, and the nonrural share is 100 minus that, so each bar always totals
100.

To see whether the Alaska and Hawaii values move the figure, add them to
`exclude_states` (for example, `c("AK", "HI")`). The chunk prints the shares
with and without the excluded states.

The title's rural share of all income-eligible children is now pulled from the
data rather than hard-coded, since moving to tract rates can change it.

```{r fig2-rural-share, fig.width=10, fig.height=4.8}
# ---- Settings to edit ---------------------------------------------------------
term_label     <- "deserts with no access to licensed home-based care, center-based care, or preschool"
exclude_states <- c()        # e.g., c("AK", "HI") pending the join check
save_png       <- TRUE       # also write a high-resolution PNG for layout

# ---- Map the rural codes to labels (and stop if the coding is unexpected) ----
rural_codes <- as.character(state_rural_summary$rural)
if (!all(rural_codes %in% c("0", "1", "FALSE", "TRUE"))) {
  stop("Unexpected values in state_rural_summary$rural: ",
       paste(unique(rural_codes), collapse = ", "),
       ". Update the label mapping in the fig2-rural-share chunk.")
}

# ---- Rural share of each group ------------------------------------------------
rural_shares <- function(df) {
  df %>%
    mutate(is_rural = as.character(rural) %in% c("1", "TRUE")) %>%
    summarise(
      all_eligible = sum(total_eligible_children[is_rural])        / sum(total_eligible_children),
      no_other     = sum(eligible_in_other_zero[is_rural])         / sum(eligible_in_other_zero),
      hs_only      = sum(eligible_in_hs_any_others_zero[is_rural]) / sum(eligible_in_hs_any_others_zero)
    ) %>%
    pivot_longer(everything(), names_to = "measure", values_to = "rural_share")
}

shares_all  <- rural_shares(state_rural_summary)
shares_used <- rural_shares(state_rural_summary %>% filter(!state_abbrv %in% exclude_states))

# Print unrounded shares (in percent) so the figure can be checked by hand
shares_all %>%
  rename(rural_pct_all_states = rural_share) %>%
  left_join(shares_used %>% rename(rural_pct_in_figure = rural_share), by = "measure") %>%
  mutate(across(where(is.double), ~ round(100 * .x, 1))) %>%
  print()
if (length(exclude_states) > 0) {
  cat("Figure excludes:", paste(exclude_states, collapse = ", "), "\n")
}

# ---- Build the bars ----------------------------------------------------------
measure_labels <- c(
  all_eligible = "All income-eligible children",
  no_other     = str_wrap(paste0("Children in ", term_label), 38),
  hs_only      = "Children for whom Head Start is the\nonly licensed option"
)

fig2_data <- shares_used %>%
  mutate(
    rural_pct    = round(100 * rural_share),
    nonrural_pct = 100 - rural_pct,          # each bar totals exactly 100
    measure      = factor(measure_labels[measure], levels = rev(measure_labels))
  ) %>%
  select(measure, Rural = rural_pct, Nonrural = nonrural_pct) %>%
  pivot_longer(c(Rural, Nonrural), names_to = "group", values_to = "pct") %>%
  mutate(group = factor(group, levels = c("Nonrural", "Rural"))) %>%  # rural sits at the left
  group_by(measure) %>%
  arrange(desc(group), .by_group = TRUE) %>%
  mutate(xmid = cumsum(pct) - pct / 2) %>%
  ungroup()

stopifnot(all(tapply(fig2_data$pct, fig2_data$measure, sum) == 100))

rural_pct_all_eligible <- fig2_data %>%
  filter(group == "Rural", measure == measure_labels[["all_eligible"]]) %>%
  pull(pct)

fig2_colors <- c(Rural = "#0B4F6C", Nonrural = "#D5DADF")

fig2 <- ggplot(fig2_data, aes(x = pct, y = measure, fill = group)) +
  geom_col(width = 0.62, color = "white", linewidth = 0.8) +
  geom_text(aes(x = xmid, label = paste0(pct, "%"),
                color = group),
            size = 4, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(values = fig2_colors, breaks = c("Rural", "Nonrural"), name = NULL) +
  scale_color_manual(values = c(Rural = "white", Nonrural = "grey20")) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 100.5)) +
  labs(
    title    = paste0("Rural children are ", rural_pct_all_eligible,
                      " percent of income-eligible children but most of\n",
                      "those with no other licensed child care option"),
    subtitle = "Rural and nonrural shares of income-eligible children under age 5, by group",
    x = NULL, y = NULL,
    caption  = paste0(
      "Note: Estimates, rounded to the nearest percent. Licensed child care deserts are areas with at least three young children per licensed slot.\n",
      "Access reflects the proximity of licensed supply, not enrollment. Nonrural includes urban and suburban areas. ",
      "Income eligibility uses census tract poverty rates for children under 5.",
      if (length(exclude_states) > 0) paste0(" Excludes ", paste(exclude_states, collapse = ", "), ".") else "",
      "\nSource: Center for American Progress analysis of licensed child care deserts data (2025) and ",
      "American Community Survey 2019–2023 five-year estimates."
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 12.5, margin = margin(b = 4)),
    plot.subtitle      = element_text(color = "grey30", margin = margin(b = 8)),
    plot.caption       = element_text(color = "grey35", size = 8, hjust = 0,
                                      margin = margin(t = 10), lineheight = 1.1),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    axis.text.x        = element_blank(),
    axis.text.y        = element_text(size = 10.5, color = "grey15", hjust = 1),
    panel.grid         = element_blank(),
    legend.position    = "top",
    legend.justification = "left",
    legend.key.size    = unit(10, "pt"),
    plot.background    = element_rect(fill = "white", color = NA),
    plot.margin        = margin(12, 16, 12, 12)
  )

fig2

if (save_png) {
  ggsave("fig2_rural_share_by_group.png", fig2, width = 10, height = 4.8, dpi = 300, bg = "white")
}
```
