---
title: "deserts_hs_analyses_2"
author: "Evan Yi"
date: "2026-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Up

```{r}
library(tidyverse)
desert_threshold <- 0.33

# Loading csv files
poverty_data <- read_csv("poverty_total_and_under5_clean.csv")
childcare_data <- read_csv("cap_ccaccess_2025_revised_02282026.csv")
joined_rural <- read_csv("joined_rural.csv")
```

## Data cleaning

```{r}
# Cleaning the poverty data
poverty_county <- poverty_data %>%
  # Extract 5-digit county FIPS from the 11-digit GEOID
  mutate(county_fips = str_sub(GEOID, 1, 5)) %>%
  filter(!is.na(pct_poverty_under5)) %>%
  group_by(county_fips) %>%
  summarise(
    avg_pct_poverty_under5 = mean(pct_poverty_under5, na.rm = TRUE),
    .groups = "drop"
  )

# Cleaning and joining the childcare data
childcare_with_poverty <- childcare_data %>%
  # Ensure county_fips is a 5-character string with leading zeros
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  left_join(poverty_county, by = "county_fips")

# Cleaning rurality data
rural_lookup <- joined_rural %>%
  select(longitude, latitude, rural) %>%
  distinct()

# Joining rurality data
childcare_with_poverty <- childcare_with_poverty %>%
  left_join(rural_lookup, by = c("longitude", "latitude"))

# Check join quality
cat("Rows with rural status:", sum(!is.na(childcare_with_poverty$rural)), "\n")
cat("Rows missing rural status:", sum(is.na(childcare_with_poverty$rural)), "\n")

# Drop missing values and code new variables 
childcare_with_poverty <- childcare_with_poverty %>%
  filter(!is.na(adj_supply_hs), !is.na(avg_pct_poverty_under5), !is.na(rural)) %>%
  mutate(
    # Estimated eligible (poverty) children per row out of 10 synthetic children
    eligible_children = 10 * (avg_pct_poverty_under5 / 100),
    # Desert indicator for Head Start
    hs_desert = adj_supply_hs <= desert_threshold
  )
```

## Calculating eligibility

VERSION NOTE: As expected, two values were adjusted slightly due to universal filtering at the top of the analysis. 

total_eligible_children: 3854947 -> 3834565
desert_eligible_children: 3820193 -> 3799876
pct_eligible_in_desert: 99.1 -> 99.1


```{r}
hs_eligible_desert <- childcare_with_poverty %>%
  summarise(
    total_eligible_children  = sum(eligible_children),
    desert_eligible_children = sum(eligible_children[hs_desert]),
    pct_eligible_in_desert   = round((desert_eligible_children / total_eligible_children) * 100, 1)
  )

print(hs_eligible_desert)
```

## Calculating eligibility by rurality

The values seemed very high, so a follow-up analysis was conducted that looked at rurality.

VERSION NOTE: As expected, no values changed. 

```{r}
hs_eligible_desert_by_rural <- childcare_with_poverty %>%
  group_by(rural) %>%
  summarise(
    total_eligible_children  = sum(eligible_children),
    desert_eligible_children = sum(eligible_children[hs_desert]),
    pct_eligible_in_desert = round((desert_eligible_children / total_eligible_children) * 100, 1),
  )

print(hs_eligible_desert_by_rural)
```

## Calculating eligibility by rurality and degree of scarcity

The values still seemed very high even with the rurality analysis. A follow-up analysis was conducted that looked at the degree of scarcity and confirmed parameters.

VERSION NOTE: As expected, no values changed. 

```{r}
# Weighted mean and median supply by rurality
childcare_with_poverty %>%
  group_by(rural) %>%
  summarise(
    weighted_mean_supply = round(
      weighted.mean(adj_supply_hs, w = eligible_children, na.rm = TRUE), 4
    ),
    median_supply = round(median(adj_supply_hs), 4),
    .groups = "drop"
  )

# Breaking down the share of areas with some supply by rurality
childcare_with_poverty %>%
  mutate(
    supply_category = case_when(
      adj_supply_hs == 0              ~ "No supply",
      adj_supply_hs <= desert_threshold ~ "Some supply, still a desert",
      TRUE                            ~ "Not a desert"
    )
  ) %>%
  group_by(rural, supply_category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(rural) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  arrange(rural, supply_category)
```

