---
title: "Child Care Desert: Poverty & Desert Access in Rural Counties"
author: "Hailey Gibbs, Center for American Progress"
date: "May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

## Data Sources and Key Definitions

| | |
|---|---|
| `childcare_data` | Lat/long points representing synthetic families from Census; `adj_supply` measure as the deserts input |
| `forhp_df` | HRSA FORHP rural county eligibility (pre-loaded) |
| NCHS data-table | Urban-rural classification scheme; rural subtypes only (Micropolitan + Noncore) |
| ACS 2023 5-Year | Pulled via tidycensus: tract-level B17001 (poverty status by sex by age, spatial join); county-level B17001, B19013, B23025, B01003 (regression controls) |

**Key definitions:**

- **Child care desert:** `adj_supply <= 0.33`
- **Rural county:** "Fully FORHP Rural" (HRSA) AND NCHS code 5 or 6
  - NCHS Micropolitan: `urb_code == 5`
  - NCHS Noncore: `urb_code == 6`
- **Poverty tract:** ≥20% of children under 6 below poverty (ACS); threshold consistent with USDA ERS classification

---

## Setup

```{r libraries}
library(tidyverse)
library(tidycensus)
library(magrittr)
library(sf)
library(tigris)
library(scales)
library(readr)
library(readxl)
library(leaflet)
library(leaflet.extras)
library(htmltools)

options(tigris_use_cache = TRUE)

## SET PATH — needed to pull the rural data files
path <- "/Users/hgibbs/Downloads/"
```

---

## Census API Key

```{r census-api-key}
# API key is stored in .Renviron as CENSUS_API_KEY.
# Run census_api_key("YOUR_KEY", install = TRUE) once to write it there.
# Do not hardcode or share the key — load it via Sys.getenv() as below.
census_api_key(Sys.getenv("CENSUS_API_KEY"), overwrite = TRUE)
```

---

## Define ACS Variable Codes

`B17001`: Poverty Status in the Past 12 Months by Sex by Age.
Children under 6 = "Under 5 years" + "5 years", by sex.

```{r acs-variables}
# Below poverty level:
#   B17001_004  Male,   Under 5,  below poverty
#   B17001_005  Male,   5 years,  below poverty
#   B17001_018  Female, Under 5,  below poverty
#   B17001_019  Female, 5 years,  below poverty
# At or above poverty level:
#   B17001_033  Male,   Under 5,  above poverty
#   B17001_034  Male,   5 years,  above poverty
#   B17001_047  Female, Under 5,  above poverty
#   B17001_048  Female, 5 years,  above poverty

poverty_vars <- c(
  below_male_u5  = "B17001_004",
  below_male_5yr = "B17001_005",
  below_fem_u5   = "B17001_018",
  below_fem_5yr  = "B17001_019",
  above_male_u5  = "B17001_033",
  above_male_5yr = "B17001_034",
  above_fem_u5   = "B17001_047",
  above_fem_5yr  = "B17001_048"
)

control_vars <- c(
  median_income = "B19013_001",  # Median household income (dollars)
  labor_force   = "B23025_003",  # Civilian labor force
  unemployed    = "B23025_005",  # Unemployed (civilian)
  total_pop     = "B01003_001"   # Total population
)
```

---

## Load Rural Classification Data

```{r rural-data}
## Rural dataset saved in Drive; available at link in GitHub
forhp_df <- read_excel(paste0(path, "rural-health-areas-data-set.xlsx"), sheet = 3)

# HRSA FORHP: fully rural counties
rural_fips_hrsa <- forhp_df %>%
  filter(County_Eligibility == "Fully FORHP Rural") %>%
  mutate(county_fips = str_pad(as.character(FIPS_2023), width = 5, pad = "0")) %>%
  pull(county_fips)

# NCHS: load and process urban-rural classification
# Restrict to codes 5 (Micropolitan) and 6 (Noncore)
nchs_df <- read_csv(paste0(path, "data-table.csv")) %>%
  separate_wider_delim(`2023 Code`, delim = " - ",
                       names = c("urb_code", "urb_category")) %>%
  mutate(
    urb_code    = as.numeric(urb_code),
    county_fips = str_pad(as.character(Location), width = 5, pad = "0")
  ) %>%
  filter(urb_code %in% c(5, 6)) %>%
  select(county_fips, State, urb_code, urb_category)

rural_fips_nchs <- nchs_df %>% pull(county_fips)

# Intersection: counties meeting BOTH criteria.
# The NCHS classification is the binding constraint — all 1,958 NCHS rural
# counties fall within the HRSA designation. The 358-county difference
# (HRSA rural but NCHS metro) is excluded; these counties have ambiguous
# rural status across federal classification systems.
rural_fips <- intersect(rural_fips_hrsa, rural_fips_nchs)

cat("HRSA Fully FORHP Rural counties     :", length(rural_fips_hrsa), "\n") ## should read 2316
cat("NCHS Micropolitan + Noncore counties:", length(rural_fips_nchs), "\n") ## should read 1958
cat("Counties meeting both criteria       :", length(rural_fips),      "\n") ## should read 1958
```

---

## Load and Filter Child Care Data

```{r load-childcare}
childcare_data <- read.csv(
  paste0(path, "cap_ccaccess_2025_revised_02282026.csv"),
  fileEncoding = "UTF-8"
) %>%
  mutate(county_name = sapply(county_name, function(x) {
    rawToChar(as.raw(utf8ToInt(x)))
  }))

childcare_rural <- childcare_data %>%
  mutate(county_fips = str_pad(as.character(county_fips), width = 5, pad = "0")) %>%
  filter(county_fips %in% rural_fips)

cat("Rural counties in childcare_data    :", n_distinct(childcare_rural$county_fips), "\n") ## should read 1949
cat("Total data points in rural counties :", nrow(childcare_rural),                   "\n") ## should read 298504
```

---

## Pull ACS Tract-Level Poverty Data

All 50 states + DC are pulled explicitly via a hardcoded FIPS list. Deriving state FIPS from `childcare_rural` would silently exclude any state with no matched rural childcare points, creating gaps in the spatial join.

Note: FIPS codes 03, 07, 14, 43, 52 do not exist.

```{r acs-state-fips}
all_state_fips <- c(
  "01","02","04","05","06","08","09","10","11","12","13",
  "15","16","17","18","19","20","21","22","23","24","25",
  "26","27","28","29","30","31","32","33","34","35","36",
  "37","38","39","40","41","42","44","45","46","47","48",
  "49","50","51","53","54","55","56"
)
```

```{r acs-main-loop}
## Main loop — pulls tract-level ACS poverty estimates for all states
tract_list <- list()

for (s in all_state_fips) {
  cat("ACS estimates — state:", s, "\n")

  tract_list[[s]] <- get_acs(
    geography   = "tract",
    variables   = poverty_vars,
    state       = s,
    year        = 2023,
    survey      = "acs5",
    geometry    = FALSE,
    output      = "wide",
    cache_table = TRUE
  )
}

tract_data <- bind_rows(tract_list)
cat("Tracts pulled in main loop:", nrow(tract_data), "\n") ## should read 84400
```

```{r acs-recovery}
## Missing state recovery
# DC (11), New Jersey (34), and Rhode Island (44) have failed transiently in
# testing and are pulled individually if missing.

states_in_data <- unique(str_sub(tract_data$GEOID, 1, 2))
missing_states <- setdiff(all_state_fips, states_in_data)

if (length(missing_states) > 0) {
  cat("States missing after main loop — pulling individually:",
      paste(missing_states, collapse = ", "), "\n")

  recovery_list <- list()

  for (s in missing_states) {
    cat("  Recovering state:", s, "\n")
    recovery_list[[s]] <- tryCatch(
      get_acs(
        geography   = "tract",
        variables   = poverty_vars,
        state       = s,
        year        = 2023,
        survey      = "acs5",
        geometry    = FALSE,
        output      = "wide",
        cache_table = FALSE
      ),
      error = function(e) {
        cat("  ERROR on state", s, ":", conditionMessage(e), "\n")
        NULL
      }
    )
    if (!is.null(recovery_list[[s]])) {
      cat("  Tracts returned:", nrow(recovery_list[[s]]), "\n")
    }
  }

  still_missing <- names(which(map_lgl(recovery_list, is.null)))
  if (length(still_missing) > 0) {
    stop(paste(
      "Recovery failed for states:", paste(still_missing, collapse = ", "),
      "\nResolve before proceeding."
    ))
  }

  tract_data <- bind_rows(tract_data, bind_rows(recovery_list)) %>%
    distinct(GEOID, .keep_all = TRUE)
}
```

```{r acs-diagnostics}
## ACS completeness checks
states_final  <- unique(str_sub(tract_data$GEOID, 1, 2))
missing_final <- setdiff(all_state_fips, states_final)
n_duplicates  <- sum(duplicated(tract_data$GEOID))

cat(sprintf("\n── ACS Pull Diagnostics ──────────────────────────────────────────────\n"))
cat(sprintf("Total tracts pulled              : %d\n",       nrow(tract_data)))   ## should read 84400
cat(sprintf("Duplicate GEOIDs                 : %d\n",       n_duplicates))       ## should read 0
cat(sprintf("States represented               : %d / %d\n",
            length(states_final), length(all_state_fips)))                         ## should read 51 / 51

if (length(missing_final) > 0) {
  stop(paste("ACS pull still incomplete. Missing states:",
             paste(missing_final, collapse = ", ")))
}
if (n_duplicates > 0) {
  stop("Duplicate GEOIDs detected. Inspect tract_data before proceeding.")
}
```

---

## Pull Tract Geometries

Pulled separately from ACS estimates to avoid the tigris cache conflict. Per-state `tryCatch` with automatic one-time retry before hard stop.

```{r tract-geometries}
tract_geo_list <- list()

for (s in all_state_fips) {
  cat("  Geometry — state:", s, "\n")
  tract_geo_list[[s]] <- tryCatch(
    tracts(state = s, year = 2023, cb = TRUE) %>%
      select(GEOID, geometry),
    error = function(e) {
      cat("  ERROR on state", s, ":", conditionMessage(e), "\n")
      NULL
    }
  )
}

tract_geo <- bind_rows(tract_geo_list)
cat("Total tract geometries pulled:", nrow(tract_geo), "\n") ## should read 84121
```

---

## Pre-Join Coverage Check

ACS-only GEOIDs (in ACS but not in geometry) fall into known special-tract categories with zero population and no poverty data:

- `990xxx` — standard water-area tracts
- `991xxx` — Hawaii water-area variants (island geography)
- `992xxx` — coastal water-area variants
- `980xxx` — group quarters / institutional population tracts

```{r pre-join-check}
n_acs_geoids <- n_distinct(tract_data$GEOID)
n_geo_geoids <- n_distinct(tract_geo$GEOID)
n_matched    <- length(intersect(tract_data$GEOID, tract_geo$GEOID))
n_acs_only   <- length(setdiff(tract_data$GEOID, tract_geo$GEOID))
n_geo_only   <- length(setdiff(tract_geo$GEOID, tract_data$GEOID))

cat(sprintf("\n── Pre-Join Coverage Check ───────────────────────────────────────────\n"))
cat(sprintf("ACS GEOIDs                       : %d\n", n_acs_geoids)) ## should read 84400
cat(sprintf("Geometry GEOIDs                  : %d\n", n_geo_geoids)) ## should read 84121
cat(sprintf("Matched GEOIDs                   : %d\n", n_matched))    ## should read 84106
cat(sprintf("In ACS only (no geometry)        : %d\n", n_acs_only))   ## should read 294
cat(sprintf("In geometry only (no ACS)        : %d\n", n_geo_only))   ## should read 15

if (n_acs_only > 0) {
  unmatched_geoids  <- setdiff(tract_data$GEOID, tract_geo$GEOID)
  special_tract_pat <- "9[89][012]\\d{3}"
  n_special         <- sum(str_detect(unmatched_geoids, special_tract_pat))
  n_unexplained     <- n_acs_only - n_special

  cat(sprintf("  Special/water/group-quarters   : %d\n", n_special))    ## should read 294
  cat(sprintf("  Unexplained (no geometry)      : %d\n", n_unexplained)) ## should read 0

  if (n_unexplained > 0) {
    problem_geoids <- unmatched_geoids[!str_detect(unmatched_geoids, special_tract_pat)]
    cat("Unexplained unmatched GEOIDs:\n")
    print(problem_geoids)
    stop(paste(
      n_unexplained,
      "non-special ACS tracts have no matching geometry.",
      "Investigate before proceeding."
    ))
  }

  cat("All ACS-only tracts are water-area, coastal, or group quarters tracts.\n")
  cat("None carry poverty data; none will affect the spatial join. Proceeding.\n")
}
```

---

## Join Estimates to Geometry and Compute Poverty Rate

```{r tract-sf}
tract_sf_raw <- tract_geo %>%
  left_join(tract_data, by = "GEOID") %>%
  st_as_sf() %>%
  st_transform(crs = 4326)

cat(sprintf("tract_sf_raw rows                : %d\n", nrow(tract_sf_raw)))

tract_sf <- tract_sf_raw %>%
  mutate(
    children_u6_below = below_male_u5E + below_male_5yrE +
                        below_fem_u5E  + below_fem_5yrE,
    children_u6_above = above_male_u5E + above_male_5yrE +
                        above_fem_u5E  + above_fem_5yrE,
    children_u6_total = children_u6_below + children_u6_above,
    # Continuous poverty rate; NA where tract has no children under 6
    pct_poverty_u6    = if_else(
      children_u6_total > 0,
      children_u6_below / children_u6_total,
      NA_real_
    )
  ) %>%
  select(GEOID, pct_poverty_u6, children_u6_total, geometry) %>%
  st_transform(crs = 4326)

cat(sprintf("tract_sf rows                    : %d\n", nrow(tract_sf))) ## should read 84121
```

---

## Pull ACS County-Level Controls

Pulled for all states to ensure coverage for every county in `rural_fips`, including any state that may have rural counties but sparse childcare points.

```{r county-controls}
county_controls <- get_acs(
  geography = "county",
  variables = c(poverty_vars, control_vars),
  year      = 2023,
  survey    = "acs5",
  geometry  = FALSE,
  output    = "wide"
) %>%
  filter(str_sub(GEOID, 1, 2) %in% all_state_fips) %>%
  mutate(
    county_fips           = GEOID,
    children_u6_below_cty = below_male_u5E + below_male_5yrE +
                            below_fem_u5E  + below_fem_5yrE,
    children_u6_above_cty = above_male_u5E + above_male_5yrE +
                            above_fem_u5E  + above_fem_5yrE,
    children_u6_total_cty = children_u6_below_cty + children_u6_above_cty,
    pct_poverty_u6_county = if_else(
      children_u6_total_cty > 0,
      children_u6_below_cty / children_u6_total_cty,
      NA_real_
    ),
    unemployment_rate = if_else(labor_forceE > 0,
                                unemployedE / labor_forceE,
                                NA_real_),
    median_income     = median_incomeE,
    total_pop         = total_popE
  ) %>%
  select(county_fips, pct_poverty_u6_county, unemployment_rate,
         median_income, total_pop)

cat("County controls pulled:", nrow(county_controls), "counties\n") ## should read 3144
```

---

## Convert Childcare Points to sf and Spatial Join

`st_within` assigns each childcare point the poverty attributes of the census tract it falls within.

```{r spatial-join}
childcare_sf <- childcare_rural %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

childcare_with_tract <- st_join(childcare_sf, tract_sf, join = st_within)
```

---

## Classify Points: Desert Status and Poverty Tiers

20% poverty and 40% extreme poverty thresholds come from USDA poverty area measures.

```{r classify-points}
# Fixed: added is_extreme_poverty — was referenced in county aggregation
# but never defined, which would have caused an error.
childcare_analysis <- childcare_with_tract %>%
  st_drop_geometry() %>%
  mutate(
    is_desert         = adj_supply <= 0.33,
    is_poverty        = !is.na(pct_poverty_u6) & pct_poverty_u6 >= 0.20,
    is_extreme_poverty = !is.na(pct_poverty_u6) & pct_poverty_u6 >= 0.40
  )
```

---

## Diagnostic: Unmatched Points

Unmatched points are water areas or edge effects.

```{r unmatched-diagnostic}
n_unmatched <- sum(is.na(childcare_analysis$pct_poverty_u6))
cat(sprintf(
  "\nPoints not matched to a tract: %d (%.1f%% of rural points)\n",
  n_unmatched,
  100 * n_unmatched / nrow(childcare_analysis)
))
childcare_analysis %>%
  filter(is.na(pct_poverty_u6)) %>%
  count(state_name, sort = TRUE) %>%
  print()

## Expected unmatched points by state:
# state_name       n
# Alaska          84
# North Carolina  13
# Washington      13
# Florida         11
# Michigan         5
# Oregon           4
# Texas            4
# Mississippi      3
# Hawaii           2
# Louisiana        2
# Virginia         2
# California       1
# Massachusetts    1
# Minnesota        1
# Missouri         1
# North Dakota     1
# Ohio             1
# Tennessee        1
```

---

## County-Level Aggregation

```{r county-aggregation}
# Desert rates are pooled across counties (total desert points / total points)
# to avoid small-county bias. Broken out by NCHS rural type x poverty tier.
county_summary <- childcare_analysis %>%
  group_by(county_fips, state_name, state_abbrv) %>%
  summarise(
    total_pts             = n(),
    desert_pts            = sum(is_desert,                      na.rm = TRUE),
    poverty_pts           = sum(is_poverty,                     na.rm = TRUE),
    extreme_poverty_pts   = sum(is_extreme_poverty,             na.rm = TRUE),
    desert_poverty_pts    = sum(is_desert & is_poverty,         na.rm = TRUE),
    desert_extreme_pts    = sum(is_desert & is_extreme_poverty, na.rm = TRUE),
    mean_tract_poverty_u6 = mean(pct_poverty_u6,                na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_desert_all     = desert_pts / total_pts,
    pct_desert_poverty = if_else(poverty_pts > 0,
                                 desert_poverty_pts / poverty_pts,         NA_real_),
    pct_desert_extreme = if_else(extreme_poverty_pts > 0,
                                 desert_extreme_pts / extreme_poverty_pts, NA_real_)
  ) %>%
  left_join(
    nchs_df %>% select(county_fips, urb_code, urb_category),
    by = "county_fips"
  ) %>%
  left_join(county_controls, by = "county_fips")
```

---

## Summary Table

```{r summary-table}
build_summary_rows <- function(data, group_label) {
  bind_rows(
    data %>%
      summarise(
        Group         = paste(group_label, "\u2014 All children"),
        N_Counties    = n(),
        Total_Points  = sum(total_pts),
        Desert_Points = sum(desert_pts),
        Pct_In_Desert = Desert_Points / Total_Points
      ),
    data %>%
      filter(poverty_pts > 0) %>%
      summarise(
        Group         = paste(group_label, "\u2014 Poverty tracts (\u226520%)"),
        N_Counties    = n(),
        Total_Points  = sum(poverty_pts),
        Desert_Points = sum(desert_poverty_pts),
        Pct_In_Desert = Desert_Points / Total_Points
      ),
    data %>%
      filter(extreme_poverty_pts > 0) %>%
      summarise(
        Group         = paste(group_label, "\u2014 Extreme poverty tracts (\u226540%)"),
        N_Counties    = n(),
        Total_Points  = sum(extreme_poverty_pts),
        Desert_Points = sum(desert_extreme_pts),
        Pct_In_Desert = Desert_Points / Total_Points
      )
  )
}

summary_table <- bind_rows(
  build_summary_rows(county_summary,                         "All rural (NCHS 5\u20136)"),
  build_summary_rows(filter(county_summary, urb_code == 5), "Micropolitan"),
  build_summary_rows(filter(county_summary, urb_code == 6), "Noncore")
) %>%
  mutate(Pct_In_Desert = percent(Pct_In_Desert, accuracy = 0.1))

print(summary_table)

## Expected output:
# Group                                                N_Counties Total_Points Desert_Points Pct_In_Desert
# 1 All rural (NCHS 5–6) — All children                        1949       298504        192328 64.4%
# 2 All rural (NCHS 5–6) — Poverty tracts (≥20%)               1574       133530         80936 60.6%
# 3 All rural (NCHS 5–6) — Extreme poverty tracts (≥40%)       1053        56666         31785 56.1%
# 4 Micropolitan — All children                                  658       184664        111709 60.5%
# 5 Micropolitan — Poverty tracts (≥20%)                         608        80885         45284 56.0%
# 6 Micropolitan — Extreme poverty tracts (≥40%)                 475        34601         17577 50.8%
# 7 Noncore — All children                                      1291       113840         80619 70.8%
# 8 Noncore — Poverty tracts (≥20%)                              966        52645         35652 67.7%
# 9 Noncore — Extreme poverty tracts (≥40%)                      578        22065         14208 64.4%
```

---

## Regression

**Unit of analysis:** rural county (HRSA Fully FORHP + NCHS codes 5–6)  
**Outcome:** `pct_desert_all`  
**Key predictor:** `mean_tract_poverty_u6` (continuous, tract-level)  
**Controls:** `median_income`, `unemployment_rate`, `log(total_pop)`, `urb_category` (Micropolitan = reference), state fixed effects  
**Weights:** `total_pts`

`urb_category` rationale: Micropolitan and Noncore counties differ systematically in population density and market size — both of which independently shape child care supply. Controlling for rurality type isolates the poverty effect from these structural baseline differences.

Model form: `pct_desert_all` is bounded [0,1]. Weighted OLS coefficients represent percentage-point changes in desert rate per unit change in each predictor.

Coefficient interpretation:

- `mean_tract_poverty_u6`: A 10 pp increase in tract child poverty rate is associated with a [coef × 10] pp change in county desert rate, holding rurality type, income, unemployment, population, and state constant.
- `urb_categoryNoncore`: Desert rate difference for Noncore vs Micropolitan counties, holding all else constant.

```{r regression}
reg_data <- county_summary %>%
  filter(
    !is.na(pct_desert_all),
    !is.na(mean_tract_poverty_u6),
    !is.na(median_income),
    !is.na(unemployment_rate),
    !is.na(total_pop),
    total_pop > 0
  ) %>%
  mutate(
    urb_category  = factor(urb_category, levels = c("Micropolitan", "Noncore")),
    state_name    = as.factor(state_name),
    log_total_pop = log(total_pop)
  )

cat(sprintf("\nRegression sample: %d rural counties\n", nrow(reg_data))) ## should read 1948

model_ols <- lm(
  pct_desert_all ~ mean_tract_poverty_u6 +
                   median_income          +
                   unemployment_rate      +
                   log_total_pop          +
                   urb_category           +
                   state_name,
  data    = reg_data,
  weights = total_pts
)

print(summary(model_ols))

## Expected output:
## lm(formula = pct_desert_all ~ mean_tract_poverty_u6 + median_income +
##      unemployment_rate + log_total_pop + urb_category + state_name,
##      data = reg_data, weights = total_pts)
##
## Weighted Residuals:
##   Min      1Q  Median      3Q     Max
## -6.0151 -0.8441  0.1581  0.9301  6.5863
##
## Coefficients:
##                         Estimate Std. Error t value Pr(>|t|)
## (Intercept)            9.777e-01  6.345e-02  15.410  < 2e-16 ***
## mean_tract_poverty_u6 -6.397e-02  3.598e-02  -1.778 0.075599 .
## median_income          1.120e-06  3.816e-07   2.935 0.003374 **
## unemployment_rate      2.892e-01  1.678e-01   1.724 0.084916 .
## log_total_pop         -3.491e-02  5.333e-03  -6.545 7.64e-11 ***
## urb_categoryNoncore    8.211e-02  7.609e-03  10.792  < 2e-16 ***
##
## Alaska                 6.688e-02  4.238e-02   1.578 0.114753
## Arizona                1.151e-01  3.573e-02   3.221 0.001299 **
## Arkansas              -6.211e-02  2.371e-02  -2.620 0.008874 **
## California            -6.240e-02  2.803e-02  -2.226 0.026121 *
## Colorado              -1.240e-01  3.001e-02  -4.132 3.76e-05 ***
## Connecticut           -8.735e-02  5.012e-02  -1.743 0.081568 .
## Delaware              -1.797e-01  4.413e-02  -4.072 4.85e-05 ***
## Florida               -7.372e-02  2.874e-02  -2.565 0.010407 *
## Georgia               -6.563e-02  2.173e-02  -3.020 0.002563 **
## Hawaii                 1.087e-01  4.052e-02   2.682 0.007377 **
## Idaho                  8.622e-02  3.091e-02   2.790 0.005331 **
## Illinois              -1.089e-01  2.268e-02  -4.804 1.68e-06 ***
## Indiana                5.870e-02  2.275e-02   2.580 0.009945 **
## Iowa                  -2.229e-01  2.435e-02  -9.155  < 2e-16 ***
## Kansas                 9.653e-03  2.629e-02   0.367 0.713529
## Kentucky              -4.800e-02  2.197e-02  -2.185 0.029020 *
## Louisiana             -4.020e-02  2.622e-02  -1.533 0.125371
## Maine                 -8.273e-02  3.244e-02  -2.550 0.010853 *
## Maryland              -1.861e-01  4.259e-02  -4.369 1.31e-05 ***
## Massachusetts         -2.512e-01  6.926e-02  -3.627 0.000294 ***
## Michigan              -6.678e-02  2.294e-02  -2.912 0.003639 **
## Minnesota             -2.106e-01  2.435e-02  -8.650  < 2e-16 ***
## Mississippi           -9.793e-02  2.218e-02  -4.416 1.06e-05 ***
## Missouri              -5.309e-02  2.258e-02  -2.351 0.018808 *
## Montana               -3.697e-02  3.170e-02  -1.166 0.243612
## Nebraska              -2.953e-01  2.829e-02 -10.439  < 2e-16 ***
## Nevada                 1.505e-03  4.292e-02   0.035 0.972036
## New Hampshire         -8.832e-02  3.531e-02  -2.502 0.012449 *
## New Mexico            -1.815e-01  2.741e-02  -6.623 4.58e-11 ***
## New York               1.066e-02  2.415e-02   0.441 0.659069
## North Carolina        -2.349e-02  2.134e-02  -1.101 0.271164
## North Dakota          -1.712e-01  3.518e-02  -4.868 1.22e-06 ***
## Ohio                  -8.597e-02  2.131e-02  -4.035 5.69e-05 ***
## Oklahoma              -1.368e-01  2.302e-02  -5.943 3.32e-09 ***
## Oregon                -1.310e-02  2.979e-02  -0.440 0.660246
## Pennsylvania           2.837e-02  2.282e-02   1.243 0.213894
## South Carolina        -6.669e-02  2.806e-02  -2.376 0.017590 *
## South Dakota          -1.196e-01  3.110e-02  -3.844 0.000125 ***
## Tennessee             -5.752e-03  2.293e-02  -0.251 0.801938
## Texas                 -1.483e-01  2.010e-02  -7.377 2.42e-13 ***
## Utah                  -1.037e-01  3.293e-02  -3.150 0.001657 **
## Vermont               -1.821e-01  3.748e-02  -4.858 1.28e-06 ***
## Virginia               1.706e-02  2.583e-02   0.661 0.508924
## Washington             3.946e-02  2.832e-02   1.394 0.163613
## West Virginia         -5.213e-02  2.898e-02  -1.799 0.072231 .
## Wisconsin             -2.135e-02  2.378e-02  -0.898 0.369453
## Wyoming               -1.874e-01  3.480e-02  -5.386 8.09e-08 ***
##
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
##
## Residual standard error: 1.482 on 1895 degrees of freedom
## Multiple R-squared:  0.3826,  Adjusted R-squared:  0.3657
## F-statistic: 22.59 on 52 and 1895 DF,  p-value: < 2.2e-16
```

---

## Interactive Map

### Prepare Map Data

```{r map-data}
county_geo <- counties(cb = TRUE, year = 2023, resolution = "5m") %>%
  filter(!STATEFP %in% c("02", "15", "72")) %>%
  mutate(GEOID = str_pad(GEOID, width = 5, pad = "0")) %>%
  select(GEOID, NAME, geometry)

map_df <- county_geo %>%
  left_join(county_summary, by = c("GEOID" = "county_fips")) %>%
  mutate(
    pct_desert_all     = if_else(GEOID %in% rural_fips, pct_desert_all,     NA_real_),
    pct_desert_poverty = if_else(GEOID %in% rural_fips, pct_desert_poverty, NA_real_)
  )

# Leaflet requires WGS84
map_df_wgs <- map_df %>%
  st_transform(crs = 4326) %>%
  mutate(
    county_label = paste0("<b>", NAME, "<b>", state_name, "</b>")
  )
```

### Build Map

```{r interactive-map}
# Color palette
pal <- colorNumeric(
  palette  = "magma",
  domain   = c(0, 1),
  na.color = "#EBEBEB"
)

# Tooltip helper
make_tooltip <- function(label, value, note = NULL) {
  pct  <- if_else(is.na(value), "No data (non-rural)", percent(value, accuracy = 0.1))
  html <- paste0(label, "<br/>Licensed child care desert rate: <b>", pct, "</b>")
  if (!is.null(note)) html <- paste0(html, "<br/><i>", note, "</i>")
  lapply(html, htmltools::HTML)
}

# Fixed: legend and addLayersControl are now piped onto interactive_map.
# In the original script they were added to a separate bare leaflet() call,
# which would have rendered the legend and layer toggle on a blank map.
interactive_map <- leaflet(
  map_df_wgs,
  options = leafletOptions(minZoom = 3)
) %>%

  # Base tile
  addProviderTiles(
    providers$CartoDB.Positron,
    options = tileOptions(opacity = 0.9)
  ) %>%

  setView(lng = -96, lat = 38, zoom = 4) %>%

  ## Layer 1: all rural children
  addPolygons(
    fillColor   = ~pal(pct_desert_all),
    fillOpacity = 0.85,
    color       = "white",
    weight      = 0.3,
    opacity     = 0.5,
    label       = make_tooltip(
      map_df_wgs$county_label,
      map_df_wgs$pct_desert_all
    ),
    labelOptions = labelOptions(
      style     = list("font-size" = "13px", "padding" = "6px 10px"),
      direction = "auto",
      textsize  = "13px"
    ),
    highlightOptions = highlightOptions(
      color        = "#333333",
      weight       = 1.5,
      fillOpacity  = 1,
      bringToFront = TRUE
    ),
    group = "All Rural Children"
  ) %>%

  ## Layer 2: poverty tracts ≥20%
  addPolygons(
    fillColor   = ~pal(pct_desert_poverty),
    fillOpacity = 0.85,
    color       = "white",
    weight      = 0.3,
    opacity     = 0.5,
    label       = make_tooltip(
      map_df_wgs$county_label,
      map_df_wgs$pct_desert_poverty,
      note = "Children in tracts \u226520% child poverty"
    ),
    labelOptions = labelOptions(
      style     = list("font-size" = "13px", "padding" = "6px 10px"),
      direction = "auto"
    ),
    highlightOptions = highlightOptions(
      color        = "#333333",
      weight       = 1.5,
      fillOpacity  = 1,
      bringToFront = TRUE
    ),
    group = "Poverty Tracts (\u226520%)"
  ) %>%

  ## Legend
  addLegend(
    position  = "bottomright",
    pal       = pal,
    values    = c(0, 1),
    title     = "% Children<br/>in Desert",
    labFormat = labelFormat(
      suffix    = "%",
      transform = function(x) round(x * 100)
    ),
    opacity = 0.85
  ) %>%

  # Layer toggle (radio buttons — only one layer visible at a time)
  addLayersControl(
    baseGroups = c(
      "All Rural Children",
      "Rural Children in Poverty (\u226520%)"
    ),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%

  ## Caption as a map control
  addControl(
    html = paste0(
      "<div style='font-size:10px; color:#666; max-width:460px; ",
      "background:rgba(255,255,255,0.85); padding:5px 8px; ",
      "border-radius:4px; line-height:1.4;'>",
      "Rural: HRSA Fully FORHP Rural + NCHS (5) and (6). ",
      "Non-rural counties in light grey.<br/>",
      "Sources: Licensed Child Care Desert Data, on file with the Center for American Progress, HRSA, NCHS, ACS 2023 5-Year Estimates.",
      "</div>"
    ),
    position = "bottomleft"
  )

## Display
interactive_map
```

### Export Map as Standalone HTML

```{r save-map, eval=FALSE}
htmlwidgets::saveWidget(
  interactive_map,
  file          = paste0(path, "interactive_desert_map.html"),
  selfcontained = TRUE
)
```
