---
title: "Child Care Deserts by Congressional District"
author: "Hailey Gibbs"
date: "April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup

Load the child care data and required libraries.

```{r load-data}
# Load child care data: CCD full file
childcare_data <- read.csv("cap_ccaccess_2025_revised_02282026.csv") # saved in Drive, linked in GitHub
```

```{r load-libraries}
library(tidyverse)
library(sf)
library(tigris)

options(tigris_use_cache = TRUE)
```

## Download and Prepare Congressional District Shapefile

```{r download-cd-shapes}
cd_shapes_119 <- congressional_districts(cb = TRUE, year = 2024)

# Check what tigris returned
attr(cd_shapes_119, "tigris")  # confirms source as CDs
names(cd_shapes_119)
nrow(cd_shapes_119)            # should be 435 + territories (441)
```

```{r filter-cd-shapes}
# Filter to 50 states + DC (exclude territories)
cd_shapes_119_filtered <- cd_shapes_119 %>%
  filter(STATEFP %in% c(sprintf("%02d", c(1:2, 4:6, 8:13, 15:42, 44:51, 53:56)), "11"))

nrow(cd_shapes_119_filtered)   # should be 436 (50 states + DC)

# Transform to WGS84
cd_shapes <- st_transform(cd_shapes_119_filtered, crs = 4326)
```

## Spatial Join: Child Care Points to Congressional Districts

```{r convert-to-sf}
# Convert child care data to SF object
df_sf <- childcare_data %>%
  filter(state_abbrv != "PR") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

```{r spatial-join}
# Join child care points to CD shapes
df_joined <- st_join(df_sf, cd_shapes[, c("GEOID", "NAMELSAD", "STATEFP", "CD119FP")],
                     join = st_within)

# Check CRS alignment and count unmatched points
st_crs(df_sf) == st_crs(cd_shapes)   # should return TRUE
sum(is.na(df_joined$GEOID))         # should be ~600 -- these are 0.027% of total data
```

### Handle Unmatched Points (Nearest Neighbor Join)

```{r fix-unmatched}
# Isolate unmatched points
unmatched <- df_joined %>% filter(is.na(GEOID))

# Find and assign nearest CD for each unmatched point
nearest_idx <- st_nearest_feature(unmatched, cd_shapes)

unmatched_fixed <- unmatched %>%
  mutate(
    GEOID    = cd_shapes$GEOID[nearest_idx],
    NAMELSAD = cd_shapes$NAMELSAD[nearest_idx],
    STATEFP  = cd_shapes$STATEFP[nearest_idx],
    CD118FP  = cd_shapes$CD118FP[nearest_idx]
  )

# Recombine with matched points
df_joined_clean <- df_joined %>%
  filter(!is.na(GEOID)) %>%
  bind_rows(unmatched_fixed)

# Verify no NAs remain
sum(is.na(df_joined_clean$GEOID))   # should be 0
```

## Calculate Child Care Deserts by Congressional District

```{r calculate-deserts}
cd_desert_estimates <- df_joined_clean %>%
  st_drop_geometry() %>%
  mutate(is_desert = adj_supply <= 0.33) %>%
  group_by(GEOID, NAMELSAD, STATEFP) %>%
  summarise(
    total_children  = n() * 10,
    desert_children = sum(is_desert) * 10,
    pct_in_desert   = desert_children / total_children * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_in_desert))

# Sanity checks
nrow(cd_desert_estimates)
summary(cd_desert_estimates$pct_in_desert)
head(cd_desert_estimates, 10)   # highest desert districts
tail(cd_desert_estimates, 10)   # lowest desert districts
```

### Remove Puerto Rico (if present)

```{r remove-pr}
# Check for duplicate GEOIDs
cd_desert_estimates %>%
  count(GEOID) %>%
  filter(n > 1)

# Check for unexpected state FIPS codes
sort(unique(cd_desert_estimates$STATEFP))

# Remove PR (STATEFP = "72")
cd_desert_estimates <- cd_desert_estimates %>%
  filter(STATEFP != "72")

# Verify
nrow(cd_desert_estimates)                    # should be 436
sort(unique(cd_desert_estimates$STATEFP))    # confirm 72 is gone
```

### Add State Abbreviations

```{r add-state-abbrvs}
state_lookup <- childcare_data %>%
  distinct(state_fips, state_abbrv) %>%
  mutate(STATEFP = sprintf("%02d", state_fips))  # pad to match STATEFP format

cd_desert_estimates <- cd_desert_estimates %>%
  left_join(state_lookup, by = "STATEFP") %>%
  select(GEOID, NAMELSAD, state_abbrv, total_children, desert_children, pct_in_desert) %>%
  arrange(desc(pct_in_desert))
```

```{r export-estimates}
write.csv(cd_desert_estimates, "cd_desert_estimates.csv", row.names = FALSE)
```

## District-Level Desert Classification

What percentage of congressional districts qualify as child care deserts?

```{r classify-districts}
DENOM_CDS <- 436

# Flag NAs before classification
cd_na_check <- cd_desert_estimates %>%
  filter(is.na(pct_in_desert))
n_na <- nrow(cd_na_check)

# Classify by threshold
cd_classified <- cd_desert_estimates %>%
  filter(!is.na(pct_in_desert)) %>%
  mutate(
    desert_50 = pct_in_desert >= 50,   # TRUE if >=50% of children in desert tracts
    desert_80 = pct_in_desert >= 80    # TRUE if >=80% of children in desert tracts
  )
```

### National Summary

```{r national-summary}
national_summary <- tibble(
  threshold          = c(">=50%", ">=80%"),
  n_desert_districts = c(sum(cd_classified$desert_50),
                         sum(cd_classified$desert_80)),
  n_non_desert       = c(sum(!cd_classified$desert_50),
                         sum(!cd_classified$desert_80)),
  n_na_districts     = n_na,
  denominator_436    = DENOM_CDS,
  pct_of_all_cds     = round(n_desert_districts / DENOM_CDS * 100, 1)
)

print(national_summary)
```

### State-Level Breakdown

```{r state-summary}
state_summary <- cd_classified %>%
  group_by(state_abbrv) %>%
  summarise(
    total_cds_in_state  = n(),
    n_desert_50         = sum(desert_50),
    n_desert_80         = sum(desert_80),
    pct_cds_desert_50   = round(n_desert_50 / total_cds_in_state * 100, 1),
    pct_cds_desert_80   = round(n_desert_80 / total_cds_in_state * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_cds_desert_50))

print(state_summary)
```

```{r export-summaries}
write.csv(national_summary, "cd_desert_national_summary.csv", row.names = FALSE)
write.csv(state_summary,    "cd_desert_state_summary.csv",    row.names = FALSE)
write.csv(cd_classified,    "cd_desert_classified.csv",       row.names = FALSE)
```

## Choropleth Map by Congressional District

### Static Map (ggplot2) (didn't love this, turned out tiny and hard to read)

```{r static-map}
library(ggplot2)

# Join estimates to spatial shapes
cd_map <- cd_shapes_119_filtered %>%
  left_join(cd_desert_estimates, by = "GEOID")

ggplot(cd_map) +
  geom_sf(aes(fill = pct_in_desert), color = NA) +
  scale_fill_viridis_c(
    option    = "magma",
    direction = -1,
    name      = "% Children\nin Desert",
    limits    = c(0, 100),
    breaks    = seq(0, 100, by = 25)
  ) +
  labs(
    title    = "Children Under 6 Living in Child Care Deserts",
    subtitle = "By Congressional District (119th Congress)",
    caption  = "A child care desert is defined as adj_supply ≤ 0.33"
  ) +
  theme_void() +
  theme(
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 10, hjust = 0.5),
    plot.caption     = element_text(size = 8, hjust = 0.5),
    legend.position  = "bottom",
    legend.key.width = unit(2, "cm")
  )

ggsave("cd_desert_map.png", width = 12, height = 7, dpi = 300)
```

### Interactive Map (Leaflet) with Tooltips (way better, preferable; doesn't sync cleanly with Datawrapper though, as a flag)

```{r leaflet-setup}
library(leaflet)
library(leaflet.extras)
library(leaflegend)

# Join estimates back to spatial shapes
cd_map <- cd_shapes_119_filtered %>%
  left_join(
    cd_desert_estimates %>% select(-NAMELSAD),
    by = "GEOID"
  ) %>%
  filter(STATEFP != "72")

# Define color palette
pal <- colorNumeric(
  palette = "magma",
  domain  = c(0, 100),
  reverse = TRUE
)
```

```{r build-tooltips}
# Build tooltip HTML for each district
cd_map <- cd_map %>%
  mutate(
    tooltip_html = ifelse(
      is.na(pct_in_desert),
      paste0(
        "<div style='font-family:\"Segoe UI\", sans-serif; padding:4px 2px;'>",
        "<b style='font-size:13px;'>", NAMELSAD, ", ", state_abbrv, "</b><br/>",
        "<span style='font-size:11px; color:#888;'>No data available</span>",
        "</div>"
      ),
      paste0(
        "<div style='font-family:\"Segoe UI\", sans-serif; padding:4px 2px; min-width:200px;'>",
        "<b style='font-size:14px;'>", NAMELSAD, ", ", state_abbrv, "</b>",
        "<hr style='margin:5px 0; border:none; border-top:1px solid #ddd;'/>",
        "<div style='margin-bottom:6px;'>",
        "<span style='font-size:11px; color:#555; text-transform:uppercase;
                      letter-spacing:0.5px;'>In a Licensed Child Care Desert</span><br/>",
        "<span style='font-size:22px; font-weight:700; color:",
        ifelse(pct_in_desert >= 66, "#b2182b",
               ifelse(pct_in_desert >= 33, "#e08214", "#2166ac")),
        ";'>", sprintf("%.1f%%", pct_in_desert), "</span>",
        "</div>",
        "<table style='width:100%; font-size:12px; border-collapse:collapse;'>",
        "<tr style='background:#f9f9f9;'>",
        "<td style='padding:3px 6px; color:#444;'>Total children represented</td>",
        "<td style='padding:3px 6px; text-align:right; font-weight:600;'>",
        scales::comma(total_children), "</td></tr>",
        "<tr>",
        "<td style='padding:3px 6px; color:#444;'>Children in desert</td>",
        "<td style='padding:3px 6px; text-align:right; font-weight:600;'>",
        scales::comma(desert_children), "</td></tr>",
        "</table></div>"
      )
    )
  )

# Verify
names(cd_map)
nrow(cd_map)  # should be 436

# Re-project to WGS84
cd_map <- st_transform(cd_map, crs = 4326)
```

```{r leaflet-map}
leaflet(cd_map) %>%
  setView(lng = -96, lat = 37.8, zoom = 4) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor    = ~pal(pct_in_desert),
    fillOpacity  = 0.8,
    color        = "white",
    weight       = 0.5,
    opacity      = 1,
    label        = lapply(cd_map$tooltip_html, htmltools::HTML),
    labelOptions = labelOptions(
      style     = list("font-weight" = "normal", padding = "3px 8px"),
      textsize  = "13px",
      direction = "auto"
    ),
    highlightOptions = highlightOptions(
      weight       = 2,
      color        = "#666",
      fillOpacity  = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegendNumeric(
    pal          = pal,
    values       = ~pct_in_desert,
    title        = htmltools::tags$div(
      style = "font-size:12px; text-align:center; font-weight:bold; margin-bottom:3px;",
      "% Young Children in a Licensed Child Care Desert"
    ),
    position     = "bottomright",
    orientation  = "horizontal",
    width        = 200,
    height       = 15,
    bins         = 5,
    numberFormat = function(x) paste0(round(x), "%")
  )
```

## Export Interactive Map

```{r export-map}
library(htmlwidgets)

map <- leaflet(cd_map) %>%
  setView(lng = -96, lat = 37.8, zoom = 4) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor    = ~pal(pct_in_desert),
    fillOpacity  = 0.8,
    color        = "white",
    weight       = 0.5,
    opacity      = 1,
    label        = lapply(cd_map$tooltip_html, htmltools::HTML),
    labelOptions = labelOptions(
      style     = list("font-weight" = "normal", padding = "3px 8px"),
      textsize  = "13px",
      direction = "auto"
    ),
    highlightOptions = highlightOptions(
      weight       = 2,
      color        = "#666",
      fillOpacity  = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegendNumeric(
    pal          = pal,
    values       = ~pct_in_desert,
    title        = htmltools::tags$div(
      style = "font-size:12px; text-align:center; font-weight:bold; margin-bottom:3px;",
      "% Young Children in a Licensed Child Care Desert"
    ),
    position     = "bottomright",
    orientation  = "horizontal",
    width        = 200,
    height       = 15,
    bins         = 5,
    numberFormat = function(x) paste0(round(x), "%")
  )

# Export as self-contained HTML
saveWidget(map, "cd_desert_map.html", selfcontained = TRUE)
```
